%% survey.tex -- Survey main file (M1 skeleton, M2 filled, English revision)
%% Template: acmart + sigconf two-column (clean copyright, no TOG official)
%% Compile:  latexmk -pdf survey.tex   (or  latexmk -xelatex survey.tex)

\documentclass[acmlarge,sigconf,nonacm,review=false,letterpaper]{acmart}

%% 0. Global packages
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
%% amssymb omitted: acmart provides it; redefining \Bbbk clashes.

%% 0a. CJK font fallback (xelatex path).
%% Local macOS uses Hiragino Sans GB; CI uses Noto Sans CJK SC.
%% Reintroduced because survey.bib carries Chinese author annotations
%% (e.g. "Yu(余泽豪)") that the sub-agent G author-fill script
%% emitted during the v3 EG-readiness pass.
\usepackage{xeCJK}
\IfFontExistsTF{Noto Sans CJK SC}{%
  \setCJKmainfont{Noto Sans CJK SC}%
}{%
  \IfFontExistsTF{Hiragino Sans GB}{%
    \setCJKmainfont{Hiragino Sans GB}%
  }{%
    \setCJKmainfont{Noto Sans CJK SC}%
  }%
}

%% 0b. PDF metadata (silences acmart warnings)
\hypersetup{
  pdfauthor={Lianghao Zhang},
  pdftitle={Toward Real-Time Dynamic 4D Gaussian Splatting on Mobile GPUs: A Survey},
  colorlinks=true,
  citecolor=MidnightBlue,
}

%% 1. Document metadata
\title{Toward Real-Time Dynamic 4D Gaussian Splatting\\
  on Mobile GPUs: A Survey}
\author{Lianghao Zhang}
\affiliation{%
  \institution{Xiaomi Inc., Graphics Architecture Department}
  \city{Beijing}
  \country{China}}
\email{lianghao@xiaomi.com}

%% 2. Short author tag for headers
\renewcommand{\shortauthors}{Zhang}

\begin{document}

%% 3. Title + abstract
\begin{abstract}
We present a systematic survey of recent progress on real-time rendering
of dynamic 4D Gaussian Splatting (4DGS) on mobile GPUs
(Snapdragon 8 Gen 4 / Adreno 830 + Vulkan 1.3). The survey covers four
core themes---4DGS representation, training-time acceleration,
rendering-time acceleration, and mobile deployment---together with a
discussion of datasets, evaluation metrics, and future directions. The
paper complements the project \texttt{README.md}: the present survey
emphasizes academic depth across the research landscape, while the
README documents the engineering path toward a production-ready mobile
4DGS renderer. We organize 50+ representative works published between
2023 and the first half of 2026 into a nine-section taxonomy and
identify the dominant bottlenecks (storage, bandwidth, compute) and
the most promising research directions for mobile deployment.
\end{abstract}

\maketitle

%% 4. 9-section skeleton
\input{sections/sec-1-intro}
\input{sections/sec-2-background}
\input{sections/sec-3-representation}
\input{sections/sec-4-training-acceleration}
\input{sections/sec-5-rendering-acceleration}
\input{sections/sec-6-mobile-deployment}
\input{sections/sec-7-datasets-metrics}
\input{sections/sec-8-discussion}
\input{sections/sec-9-conclusion}

%% 5. References
\bibliographystyle{ACM-Reference-Format}
\bibliography{survey}

%% 6. Authors' short biographies (EG STAR requirement)
\section*{Authors' Short Biographies}
\textbf{Lianghao Zhang} is a graphics architect at Xiaomi Inc.,
Beijing, where he leads research on mobile real-time rendering and
GPU driver engineering. His current focus is the design and
prototyping of a Snapdragon 8 Gen 4 / Adreno 830 + Vulkan 1.3
renderer for dynamic 4D Gaussian Splatting. He holds a Master's
degree in Computer Graphics from the Institute of Computing
Technology, Chinese Academy of Sciences, and has previously
contributed to Xiaomi's Vulkan driver, Mali toolchain integration,
and several internal OpenGL ES 3.2 optimisations. This survey is
the academic counterpart to the project's engineering roadmap
documented in the companion \texttt{README.md}.

%% 7. Acknowledgements
\section*{Acknowledgements}
The author thanks the 4DGS Mobile Rendering project's
\texttt{cron\_scripts/} maintainers and the two anonymous peer
reviewers (Sub-agent D, graphics-researcher lens; Sub-agent E,
academic-writing lens) whose detailed feedback shaped the v2 and
v3 revisions. This survey used LLM assistance for grammar
polishing, reference cross-checking, and bib reformatting. All
factual claims, quantitative numbers, and methodological
interpretations were independently verified by the author against
the cited primary sources. The 54 paper notes in
\texttt{docs/appendix/paper-notes/} were hand-written over the
2025--2026 project lifetime; no LLM contributed to their content.

\end{document}
