%% sec-8-discussion.tex -- Discussion & Future Directions

\section{Discussion and Future Directions}
\label{sec:discussion}

\subsection{Five Promising Directions}

\begin{enumerate}
  \item \textbf{Feed-forward 4DGS.} Eliminate per-scene
        optimisation. L2D2-GS~\cite{song2026l2d2gs} opens this path;
        we expect end-to-end ``video in, 4DGS out'' models in the
        near future.
        \begin{itemize}
        \item \textit{Missing piece:} an open, multi-view dynamic
              dataset large enough to train a feed-forward 4DGS
              backbone comparable in scale to Objaverse-XL.
        \item \textit{What the next paper should look like:} a
              transformer-based densifier that consumes monocular
              video at 30~FPS and emits a 4DGS scene in one pass,
              evaluated on DyCheck~\cite{gao2022dycheck}.
        \item \textit{Project fit:} \checkmark{}~feed-forward maps
              directly to Snapdragon~8~Elite inference throughput.
        \item \textit{Quantitative target:} by 2027~H1, a feed-forward
              4DGS backbone that reconstructs a 10-second DyCheck clip
              in $<$30~s on a single Adreno~830, matching the
              per-scene optimisation quality of Deformable-3DGS
              (LPIPS $\le 0.18$) within $5\%$.
        \end{itemize}

  \item \textbf{Hardware-software co-design.} Neo~\cite{oh2025neo}'s
        Reuse-and-Update sorting accelerator hints at dedicated
        4DGS compute units inside mobile GPUs.
        \begin{itemize}
        \item \textit{Missing piece:} a public 4DGS workload
              characterisation (FLOPS, tile occupancy, register
              pressure) on Adreno-class mobile GPUs.
        \item \textit{What the next paper should look like:} a
              tile-sort accelerator IP synthesised for a mobile
              shader core, benchmarked against the
              Vulkan~1.3 compute path.
        \item \textit{Project fit:} \checkmark{}~the Adreno~830
              shader ISA and Vulkan~1.3 compute-shader model are
              the deployment substrate.
        \item \textit{Quantitative target:} a co-designed sort
              accelerator that reduces the per-Gaussian sort cost
              by $3\times$ versus the software-only Vulkan~1.3
              baseline, measured on the Mip-NeRF~360 bicycle
              scene at 1080p.
        \end{itemize}

  \item \textbf{4DGS ray tracing.} The 4DGRT~\cite{liu20254dgrt}
        route is the long-term path to high fidelity, but mobile ray
        tracing remains an open problem.
        \begin{itemize}
        \item \textit{Missing piece:} mobile RT cores are
              bandwidth-starved for the per-Gaussian intersection
              workload; no published 4DGS BVH layout.
        \item \textit{What the next paper should look like:} a
              two-level BVH that exploits Gaussian covariance
              locality, evaluated against the rasteriser baseline
              on Snapdragon~8~Elite.
        \item \textit{Project fit:} \textbf{!}~mobile RT is not
              on the immediate Snapdragon~8~Elite / Vulkan~1.3
              hot path; keep as a 2027+ watch item.
        \item \textit{Quantitative target:} by 2028, a BVH-based
              4DGS ray tracer that reaches 15~FPS @ 720p on
              a future Snapdragon (Gen~5 or later) mobile GPU,
              within $2\times$ of the rasteriser FPS.
        \end{itemize}

  \item \textbf{Temporal compression.} Dynamic / static separation
        + residual coding + contextual coding may converge into a
        4DGS compression standard.
        \begin{itemize}
        \item \textit{Missing piece:} an MPEG-style codec profile
              that standardises the dynamic / static split, the
              residual format, and the rate-control interface.
        \item \textit{What the next paper should look like:} a
              reference encoder / decoder pair with bitstream
              conformance tests, integrating
              4DGS-CC~\cite{chen20254dgscc} and
              P-4DGS~\cite{wang2025p4dgs}.
        \item \textit{Project fit:} \checkmark{}~a codec standard
              directly benefits our project's storage bottleneck.
        \item \textit{Quantitative target:} a reference codec that
              compresses the 4DGS-1K D-NeRF scenes to $<$20~MB
              per scene at PSNR $\ge 30$~dB, with a
              decoder-side decode time under 50~ms on
              Adreno~830.
        \end{itemize}

  \item \textbf{Cross-domain fusion.} 4DGS + physical simulation
        (GaussianFluent~\cite{huang2026gaussianfluent}), 4DGS + SLAM,
        and 4DGS + on-device large multimodal models are all
        emerging fronts.
        \begin{itemize}
        \item \textit{Missing piece:} no end-to-end system paper
              that closes the loop between 4DGS reconstruction,
              physics simulation, and on-device LLM/VLM grounding.
        \item \textit{What the next paper should look like:} a
              demo of ``video in, simulator-in-the-loop 4DGS out''
              running in real time on a mobile device.
        \item \textit{Project fit:} \textbf{!}~valuable but
              orthogonal to the core rendering acceleration goal;
              potential follow-up after the codec-standard work.
        \item \textit{Quantitative target:} a cross-domain demo
              that combines 4DGS reconstruction with a
              fracture-aware simulation and renders the result at
              $\ge 10$~FPS on a mobile device.
        \end{itemize}

  \item \textbf{Generative 4DGS.} The L2D2-GS feed-forward route
        (\S\ref{sec:training}) generalises to a wider class of
        generative 4DGS models. The natural extension is
        text-to-4DGS (analogous to DreamFusion's text-to-3D path)
        and image-to-4DGS (single-image dynamic scene
        reconstruction).
        \begin{itemize}
        \item \textit{Missing piece:} a public, large-scale
              text-paired dynamic video corpus (comparable in
              scale to Objaverse-XL or LAION-5B for static
              3D generation).
        \item \textit{What the next paper should look like:} a
              diffusion-based 4DGS generator that consumes a
              text prompt + a single static 3DGS seed and
              outputs a 4DGS scene in $<$10~seconds on an
              Adreno~830, evaluated on a held-out subset of
              DyCheck prompts.
        \item \textit{Project fit:} \textbf{!}~exciting but
              currently compute-bound; a Snapdragon~8~Elite
              target is plausible only after the mobile
              transformer cost analysis matures.
        \item \textit{Quantitative target:} by 2027~H2, a
              text-to-4DGS system that produces a 10-second
              dynamic clip on a single Adreno~830 in
              $<$30~seconds, with LPIPS $\le 0.20$ against the
              ground-truth clip.
        \end{itemize}

  \item \textbf{4DGS standardisation.} The 4DGS community
        currently has no standard codec profile, no standard
        benchmark suite, and no standard mobile evaluation
        harness. Three pieces of standardisation are likely
        in the next two years.
        \begin{itemize}
        \item \textit{Missing piece:} an MPEG- or
              Khronos-style codec profile that standardises the
              dynamic / static split, the residual format, and
              the rate--control interface. The
              4DGS-CC~\cite{chen20254dgscc} + P-4DGS~\cite{wang2025p4dgs}
              + PD-4DGS trio are the de-facto building
              blocks.
        \item \textit{What the next paper should look like:} a
              reference encoder / decoder pair with bitstream
              conformance tests, distributed as an open-source
              reference implementation.
        \item \textit{Project fit:} \checkmark{}~the
              reference codec is the natural first deliverable
              of the project's engineering roadmap
              (see \S\ref{sec:mobile}).
        \item \textit{Quantitative target:} a reference
              decoder that runs at 30~FPS sustained on a
              Snapdragon~8~Elite device, on a 20~MB 4DGS
              bitstream, with PSNR $\ge 30$~dB.
        \end{itemize}
\end{enumerate}

\subsection{Why Hardware--Software Co-Design Will Dominate (Extended)}
We expect \emph{hardware--software co-design} to matter more than
algorithmic progress for mobile 4DGS over the next two years. The
argument has three legs.

\paragraph{First, the algorithmic headroom is shrinking.}
The 4DGS literature has seen roughly 3$\times$ per-year gains in
training speed (4DGS 30--60~min in 2023, OMG4 22~min in 2025,
L2D2-GS $<$1~s in 2026 at inference) and 5--10$\times$ per-year
gains in model size (vanilla 4DGS 60--200~MB in 2023,
4DGS-1K $\sim$7--418~MB in 2025 (D-NeRF 7--16~MB, N3V 41--418~MB),
4DGS-CC 1.7--5.0~MB in 2025). Both
curves are flattening: the next 3$\times$ in either dimension
will require a fundamentally new representation (e.g. generative
4DGS, see \S8.6 above), not an incremental tweak of the existing
one.

\paragraph{Second, the hardware headroom is large and quantifiable.}
The Adreno~830's tile-sort cost (10--20\% of the frame budget at
60~FPS, see \S\ref{sec:rendering}) is a structural artefact of
the rasteriser's $\alpha$-blend rule
(Equation~\eqref{eq:alpha}). No algorithm change can amortise that
cost; only a hardware change can. Neo's Reuse-and-Update sort
(\S\ref{sec:rendering}) is a software prototype of what a
dedicated sort accelerator would do, and the 3--5$\times$ sort
speed-up it delivers would translate directly to a
30--50\% end-to-end speed-up at the 30~FPS sustained target.

\paragraph{Third, the co-design loop is short.}
The Snapdragon mobile launch cadence is roughly 12~months
(Gen~3 $\to$ Gen~4 $\to$ Gen~5 in 2024/2025/2026), which means
the algorithm-to-silicon loop is at most 24~months. The Adreno
shader ISA is stable across generations (no breaking changes
from Gen~3 to Gen~4), and the Vulkan~1.3 driver is upstreamed
into the Linux kernel with a 6-month cadence. These three
properties together make a tight algorithm / driver / ISA
co-design loop tractable, which is the structural advantage
that mobile 4DGS has over its desktop counterpart. The
contrarian view is therefore not ``algorithm doesn't matter''
but rather ``the next 3$\times$ on mobile will come from
hardware / driver / ISA co-design, not from a better MLP''.

\subsection{Contrarian View: Hardware--Software Co-Design Will Dominate}
We expect \emph{hardware--software co-design} to matter more than
algorithmic progress for mobile 4DGS over the next two years. The
reason is structural: Vulkan driver maturity, tile-based rendering
alpha-blend cost, and the absence of a dedicated 4DGS compute unit
on Adreno~830 are first-order constraints that no MLP redesign can
bypass. In other words, even a $10\times$ better densifier will
still hit the same alpha-blend fillrate wall. The most leveraged
research investment today is therefore a tight co-design loop
between the algorithm and the Vulkan~1.3 / Adreno shader ISA,
rather than yet another MLP variant.

\subsection{Connection to This Project}
The project \texttt{README.md} fixes the target on
Snapdragon~8~Elite + Vulkan~1.3 real-time rendering, which
overlaps the five directions above. See \texttt{01-} / \texttt{02-}
/ \texttt{03-} for the engineering roadmap.

\bigskip
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S G.
