%% sec-1-intro.tex -- Introduction

\section{Introduction}
\label{sec:intro}

Real-time neural rendering of dynamic scenes is one of the central
research thrusts in computer graphics over the past few years. Since
its introduction in 2023, 3D Gaussian Splatting (3DGS)~\cite{kerbl20233dgs}
has demonstrated that an explicit point-cloud representation combined
with differentiable rasterization can deliver 1080p real-time radiance
field rendering on desktop GPUs. Extending 3DGS to dynamic scenes
(4DGS), however, raises three core challenges: (\emph{i})~the choice of
temporal parameterization; (\emph{ii})~the storage cost of separating
dynamic and static content; and (\emph{iii})~the compute, bandwidth,
and memory budget of mobile GPUs.

\paragraph{Brief history of real-time neural rendering.}
The 4DGS research programme inherits from three earlier
threads. \emph{NeRF and its descendants} (2020--2022) demonstrated
that an MLP could represent a 3D radiance field with high
quality, but the volumetric rendering loop was too slow for
real-time use. \emph{Instant-NGP and the hash-grid family}
(2022) reduced the per-sample cost by 2--3 orders of magnitude
via a multi-resolution hash encoding, but were still
constrained to $\sim$10--15~FPS at 1080p. \emph{3D Gaussian
Splatting} (2023) broke this barrier by replacing the MLP
with an explicit Gaussian set and a differentiable rasteriser,
reaching 100+~FPS at 1080p on RTX A6000 (Kerbl 2023, Table 2). The 4DGS
literature (2023--) is the next step in this line: it lifts
the 3DGS speed advantage to dynamic scenes, at the cost of a
4--5$\times$ parameter explosion that motivates the
compression, streaming, and mobile-optimisation routes
surveyed in \S\ref{sec:training}--\S\ref{sec:mobile}.

\paragraph{The 4DGS timeline (2023--H1 2026).}
The 4DGS research programme can be divided into three
generations. \emph{Gen 1 (2023--2024, canonical + deformation):}
4DGS~\cite{wu20244dgs}, Deformable-3DGS~\cite{yang2023deformable3dgs},
and the HyperNeRF-derived work establish the canonical
parameterisation and per-frame deformation MLP paradigm.
\emph{Gen 2 (2024--2025, dynamic / static separation):}
4DGS-1K~\cite{yuan20254dgs1k}, MEGA~\cite{zhang2024mega4dgsacceleration},
and 4DGS-CC~\cite{chen20254dgscc} split the Gaussian set into
a static buffer-A and a dynamic buffer-B, achieving the
first 1000+~FPS dynamic reconstruction. \emph{Gen 3 (2025--2026,
feed-forward + hardware):} L2D2-GS~\cite{song2026l2d2gs},
Mobile-GS~\cite{du2026mobilegs}, Flux-GS~\cite{du2026fluxgs},
and Neo~\cite{oh2025neo} attack the per-scene optimisation
loop and the mobile sort cost, respectively. The three
generations are not strictly sequential---a 2026 paper can
mix elements of all three---but the dominant research
question of each year follows the pattern above.

\paragraph{Contributions of this survey.}
This survey makes the following deliverable contributions, each stated
as a verifiable artefact rather than a generic claim.

\begin{description}
  \item[\textbf{C-1.}]
    A unified taxonomy that maps 4DGS research onto five
    \emph{factions} matching the rendering pipeline stages
    (representation / compression / rendering / mobile deployment /
    cross-disciplinary). Each of the 60 surveyed papers is annotated
    as \emph{3DGS-only}, \emph{4DGS-native}, or \emph{transferable},
    and the per-paper labels are exposed in Appendix~A so that
    readers can re-classify any single work without re-reading it.

  \item[\textbf{C-2.}]
    A systematic on-device FPS / model-size / energy survey across
    seven or more mobile 4DGS works with an explicit \emph{data-source}
    column. Each numeric claim is tagged \texttt{[measured]} /
    \texttt{[abstract only]} / \texttt{[projected]} so that readers can
    distinguish measurements taken from the paper text, digits reused
    from the abstract without verification, and forward projections.

  \item[\textbf{C-3.}]
    A three-walls quantification of the mobile deployment gap:
    (\emph{i})~model size above the 200~MB Google Play AAB cap;
    (\emph{ii})~cold-start above the Android~Vitals 5~s threshold;
    (\emph{iii})~sustained 720p rendering below 5~FPS on mobile
    targets. The contrast with the desktop RTX~3090 envelope of
    approximately 350~W versus the Snap~8~Elite sustained envelope
    of 5--8~W makes the 4DGS-on-mobile gap measurable rather than
    rhetorical.

  \item[\textbf{C-4.}]
    A research roadmap cross-referencing the three walls with
    concrete intervention points: Hexagon NPU offload for spherical
    harmonics evaluation, push-descriptor binding for per-frame
    dynamic uniforms, subgroup-sort for vanilla 3DGS pipelines, and
    a 5\%-tile memory bandwidth budget. Each intervention is anchored
    to a 2027 H1 / 2028 / 2030 milestone in \S\ref{sec:discussion}.

  \item[\textbf{C-5.}]
    The first public per-paper 3DGS / 4DGS transferability table
    (Appendix~A), so that future surveys can build on a structured
    corpus rather than re-doing the taxonomy work from scratch.
    The three-state label (\emph{3DGS-only} / \emph{transferable} /
    \emph{4DGS-native}) is the smallest unit that still distinguishes
    a desktop-only optimization from one that maps cleanly onto the
    4DGS mobile pipeline.
\end{description}
\noindent To the best of our knowledge, no prior survey has packaged
these five artefacts together.

This survey focuses on \emph{real-time 4DGS on mobile devices} and
systematizes more than 50 representative works published between 2023
and the first half of 2026. We organize the field around the following
guiding questions:
\begin{itemize}
  \item What 4DGS representation routes exist, and what are their
        trade-offs?
  \item How can the training stage (pruning, quantization, temporal
        redundancy) be compressed?
  \item How can the rendering stage (rasterization, ray tracing, mesh
        extraction) be accelerated?
  \item What are the current bottlenecks of Snapdragon / Adreno, and
        what are the most promising paths forward?
\end{itemize}

\paragraph{Scope and explicit boundaries (B-1 to B-10).}
To prevent the survey from being read as either over-broad or
under-measured, we state the following ten scope boundaries
up-front. Each boundary is honoured throughout \S2--\S9.
\begin{description}
  \item[\textbf{B-1.}] The survey consolidates 50+ paper notes
        (54 in practice after the per-paper labels of Appendix~A are
        applied).
  \item[\textbf{B-2.}] Approximately half of the surveyed works are
        static 3DGS acceleration that does not extend cleanly to
        4DGS; the 4DGS-native subset is below half of the corpus.
  \item[\textbf{B-3.}] \textbf{Zero} of the surveyed papers report an
        on-device measurement of dynamic 4DGS on the project target
        platform (Snapdragon~8~Elite + Adreno~830 + Vulkan~1.3).
        Every Snap~8~Elite, Adreno~830, or Vulkan~1.3 digit quoted
        in this survey is either an extrapolation from a desktop
        GPU or a Snap~8~Gen~3 baseline, or a leaked
        engineering-sample figure; none of them is a 4DGS-on-mobile
        measurement.
  \item[\textbf{B-4.}] The closest \emph{measured} mobile-class
        platforms in the corpus are Snap~8~Gen~3 (Adreno~750),
        Jetson~Orin (Ampere, 60--100~W edge envelope), and the
        mobile-Volta class. Desktop RTX~3090 and RTX~A6000 numbers
        appear only as algorithmic baselines, not as deployment
        evidence.
  \item[\textbf{B-5.}] Any numeric claim whose value was not
        verified inside the paper note is tagged
        \texttt{[abstract only]} (digit reused from the abstract
        without cross-check) or \texttt{[projected]} (forward
        estimate from a SOTA curve). Untagged digits are measured.
  \item[\textbf{B-6.}] This survey does \emph{not} replace the project
        \texttt{README.md}, nor does the project depend on this
        survey for its engineering decisions. The two artefacts
        serve distinct audiences (researchers vs.\ engineers) and
        are cross-referenced only where overlap is unavoidable.
  \item[\textbf{B-7.}] The scope is restricted to the four pipeline
        stages representation / compression / rendering / mobile
        deployment, plus the cross-disciplinary faction~E. Other
        neural-rendering pipelines (NeRF variants, hash grids, mesh
        reconstruction) appear \emph{only} as comparison baselines.
  \item[\textbf{B-8.}] Roughly 40\% of the 60 corpus entries are
        pre-prints; the \texttt{pre-print} status is flagged
        explicitly in the bibliography.
  \item[\textbf{B-9.}] The survey deliberately excludes
        6-DoF SLAM, dense 3D reconstruction from RGB-D / LiDAR, mesh
        animation, and other dynamic representations that do not
        share the 3DGS rendering loop.
  \item[\textbf{B-10.}] The cutoff is 30~June 2026; works appearing
        after this date are deferred to a successor survey and are
        not retro-fitted into the present taxonomy.
\end{description}

\noindent\textbf{Faction classification standard.}
The five factions used throughout the survey follow the rendering
pipeline axis rather than a publication-year or authorship axis:
\textbf{A = Representation} (how the dynamic Gaussian set is
parameterised); \textbf{B = Compression} (how the parameterised set
is reduced for storage or transmission); \textbf{C = Rendering}
(how the parameterised or compressed set is rasterised, ray-traced,
or otherwise projected); \textbf{D = Mobile Deployment} (how the
rendering stage is mapped onto a Snapdragon / Adreno / Vulkan
target); \textbf{E = Cross-Disciplinary} (static-3DGS acceleration
\emph{and} adjacent 2D / video / NeRF work that informs the mobile
4DGS design space but does not stand on the A--D pipeline on its
own). A paper is assigned to the single faction whose pipeline
stage is the load-bearing novelty; work that legitimately splits
across two stages is cross-listed in both Appendix~A and the
in-text discussion. Faction~E is therefore not ``miscellaneous''
--- it is the orthogonal bucket of orthogonal design aids, and
its omission would force the A--D pipeline to be re-opened every
time a static-3DGS paper is cited.

\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S A.

\bigskip
\noindent\textbf{Research-question roadmap.} The four guiding questions
are answered in the following sections:
\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Question} & \textbf{Answered in} \\
\midrule
Q1: 4DGS representation routes and trade-offs? & \S\ref{sec:representation} \\
Q2: Training-time compression routes? & \S\ref{sec:training} \\
Q3: Rendering-time acceleration routes? & \S\ref{sec:rendering} \\
Q4: Mobile deployment + future directions? & \S\ref{sec:mobile}, \S\ref{sec:datasets}, \S\ref{sec:discussion} \\
\bottomrule
\end{tabular}
\end{center}

\begin{figure}[t]
\centering
\small
\begin{tabular}{p{0.18\linewidth}p{0.72\linewidth}}
\toprule
\textbf{Faction} & \textbf{Scope and representative work} \\
\midrule
\textbf{A. 4DGS representation} &
Canonical + deformation (\cite{wu20244dgs}, \cite{yang2023deformable3dgs}); dynamic / static separation (\cite{yuan20254dgs1k}, \cite{zhang2024mega4dgsacceleration}); continuous-time (\cite{wang2026retimegs}); memory-efficient (\cite{shi2025sparse4dgs}, \cite{ren2026cubifygs}). \\
\textbf{B. 4DGS training accel.} &
Temporal pruning (\cite{lee2025omg4}, \cite{tu2025speede3dgs}); contextual / progressive coding (\cite{chen20254dgscc}, \cite{li2026pd4dgs}); streamable layered coding (\cite{yin2026cags}, \cite{zheng20254dgcpro}). \\
\textbf{C. 3DGS rendering accel.} &
Mobile rasterisation (\cite{du2026mobilegs}, \cite{du2026fluxgs}, \cite{huang2025seele}); ray tracing (\cite{liu20254dgrt}, \cite{ghosh2026gsnfs}); mesh / mobile neural rendering (\cite{feng2025lumina}); hardware acceleration (\cite{oh2025neo}). \\
\textbf{D. Mobile + streaming} &
Streaming codecs (\cite{li2026pd4dgs}, \cite{ke2025streamstgs}, \cite{veicht2026zipsplat}, \cite{yu2026codecsplat}); on-device decode (\cite{ghosh2026gsnfs}, \cite{shi2026evogs}); LoD (\cite{shi2026evogs}). \\
\textbf{E. Static 3DGS / cross-disciplinary} &
Compression primers (\cite{fan2023lightgaussian}, \cite{chen2024hacpp}, \cite{navaneet2023compact3d}); memory-efficient inference (\cite{liu2024efficientgs}, \cite{li2026vedal}); cross-domain (\cite{huang2026gaussianfluent}, \cite{feng2025lumina}). \\
\bottomrule
\end{tabular}
\caption{Faction taxonomy used throughout this survey. Factions are
ordered along the 4DGS rendering pipeline rather than along a
publication-year or authorship axis: \textbf{A = 4DGS Representation}
(how the dynamic Gaussian set is parameterised);
\textbf{B = 4DGS Compression / Training Acceleration} (how the
parameterised set is reduced for storage or transmission);
\textbf{C = Rendering Acceleration} (how the parameterised or
compressed set is rasterised, ray-traced, or otherwise projected);
\textbf{D = Mobile Deployment \& Streaming} (how the rendering stage
is mapped onto a Snapdragon / Adreno / Vulkan target);
\textbf{E = Cross-Disciplinary} (static-3DGS acceleration and
adjacent 2D / video / NeRF work that informs the mobile 4DGS design
space but does not stand on the A--D pipeline on its own). Each
surveyed paper is assigned to the single faction whose pipeline
stage is the load-bearing novelty; papers that legitimately split
across two stages are cross-listed in both Appendix~A and the
in-text discussion. The body of the survey is organised into nine
sections (\S2--\S9) following the same
representation / compression / rendering / deployment axis, with
cross-references back to the appropriate faction where needed.}
\label{fig:taxonomy}
\end{figure}
