%% sec-9-conclusion.tex -- Conclusion

\section{Conclusion}
\label{sec:conclusion}

We have systematised more than 50 representative works on
real-time dynamic 4DGS rendering on mobile GPUs published between
2023 and the first half of 2026. The survey is organised around
four themes: representation, training-time acceleration,
rendering-time acceleration, and mobile deployment. Current
Snapdragon~8~Elite devices already achieve desktop-class real-time
3DGS, but 4DGS remains limited by a triple bottleneck of memory
bandwidth, Vulkan driver maturity, and the mobile power wall.

Promising future directions include feed-forward 4DGS,
hardware--software co-design, 4DGS ray tracing, a temporal
compression standard, and cross-modal fusion. We expect the
emergence of a community-wide \emph{4DGS compression standard}---a
codec profile that combines dynamic/static separation, residual
coding, and rate--distortion optimisation; preliminary prototypes
along this line include 4DGS-CC~\cite{chen20254dgscc} and the
feed-forward densification policy of
L2D2-GS~\cite{song2026l2d2gs}, both of which could anchor such a
standard. This survey
complements the project \texttt{README.md}: the present document
emphasises academic depth across the research landscape, while the
README documents the engineering path toward a production-ready
mobile 4DGS renderer.

\subsection{Limitations of This Survey}
We acknowledge three known limitations of the present survey.
First, in \textbf{scope}: we cover 2023 to the first half of
2026; very early 4DGS works from 2022 (e.g.\ the original NeRF-t
lineage) and any submission that appears after the cut-off are
excluded. Second, in \textbf{methodology}: the survey is
English-only and prioritises code-available or reproducible
papers; non-English venues and theoretical-only submissions are
underrepresented. Third, in \textbf{known gaps}: (i)~Faction~A
(the earliest 4DGS canonical-space methods) receives less space
than the newer feed-forward and compression factions;
(ii)~feed-forward 4DGS is covered via L2D2-GS but the parallel
diffusion-based generation literature is only sampled; and
(iii)~cross-domain fusion papers (4DGS + SLAM,
4DGS + simulation) are sparse and we surface them only
in~\S\ref{sec:discussion}.

\subsection{Reproducibility and Open Science}
This survey is reproducible from the project repository. The
\texttt{survey.tex} source compiles with a single
\texttt{latexmk -pdf survey.tex} invocation (or the
\texttt{make} target in \texttt{docs/survey/Makefile}); the
required \texttt{acmart} class ships with \texttt{texlive-full}.
A GitHub Action at \texttt{.github/workflows/survey.yml}
re-compiles and republishes the PDF on every push to
\texttt{main}, with the compiled artifact attached to the run.
The bibliography in \texttt{survey.bib} is regenerated from the
54 paper notes in \texttt{docs/appendix/paper-notes/} via
\texttt{scripts/assign\_survey\_section.py} (faction $\to$
section mapping) and \texttt{scripts/extract\_paper\_summary.py}
(per-section \verb|\item| blocks and bib entries). Every
\texttt{\textbackslash cite} key in this PDF corresponds to a
paper note in the repository, which records the source PDF page
for every quantitative claim, the citation chain, and a
faction-classified summary. The reader is invited to file issues
on the project repository for missing papers, mis-categorised
factions, or updated performance numbers.

\bigskip
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md}.
