%% sec-7-datasets-metrics.tex -- Datasets & Metrics

\section{Datasets and Metrics}
\label{sec:datasets}

\subsection{Commonly Used Datasets}
4DGS training and evaluation rely on a handful of well-established
dynamic-scene datasets. We organise them along three axes:
synthetic vs.\ real, monocular vs.\ multi-view, and the
capture-rig geometry. The five datasets below cover the
\emph{de-facto} standard benchmarks; further, less-common
captures (e.g.\ the INRIA synthetic-human dataset, the
Waymo-open dynamic set, the DeepView light-field video
collection) are used by individual works but rarely as
cross-paper comparison points.

\paragraph{Synthetic datasets.}
\begin{itemize}
  \item \textbf{D-NeRF}~\cite{pumarola2021dnerf} (synthetic): eight
        synthetic monocular scenes (e.g.\ Stand~Up, Hell~Warrior,
        Bouncing~Balls) with simple rigid and articulated
        deformations; remains the standard synthetic baseline for
        4DGS benchmarks. D-NeRF provides a known ground-truth camera
        trajectory, which makes it the cleanest ablation target
        for the canonical + deformation route of
        \S\ref{sec:representation}.
  \item \textbf{DyCheck synth} (DyCheck-iPhone split): the
        synthetic DyCheck split provides known camera paths and
        per-frame mesh ground truth, which is the substrate that
        makes the dynamic / static separation route of
        \S\ref{sec:representation} measurable.
\end{itemize}

\paragraph{Real-world multi-view datasets.}
\begin{itemize}
  \item \textbf{Plenoptic Video}~\cite{li2022plenvdb} (Google): a
        12-camera array dataset (\emph{Ballroom}, \emph{Cut~Roasted
        Beef}, etc.) with high-quality multi-view captures used by
        4DGS papers that need dense temporal supervision. The
        12-camera rig provides a $360^{\circ}$ coverage and
        sub-frame synchronisation, which is the standard
        multi-view rig for 4DGS work that needs cross-frame
        Gaussian correspondence.
  \item \textbf{CMU Panoptic}~\cite{simon2017cmu}: multi-view,
        multi-person interactions captured by a 31-camera dome;
        the standard multi-person benchmark for 4DGS in crowd
        and pose scenarios. CMU Panoptic is the only
        widely-used dataset that captures the social-scene
        dynamics (5--15 simultaneous actors) that mobile
        4DGS must eventually handle.
  \item \textbf{Neu3D / Technicolor Light-Field}: a
        16-camera light-field rig that records 12-bit HDR
        frames; the highest-fidelity multi-view rig in the
        4DGS literature, but the file sizes ($\sim$5~TB per
        scene) make it impractical for routine use.
\end{itemize}

\paragraph{Real-world monocular datasets.}
\begin{itemize}
  \item \textbf{DyCheck}~\cite{gao2022dycheck} (iPhone): real-world
        dynamic scenes captured with a handheld iPhone, with
        held-out multi-view ground truth; the de-facto real-world
        benchmark for 4DGS quality evaluation. The DyCheck
        train/test split is the one most 4DGS papers use
        to report numbers.
  \item \textbf{Nerfies}~\cite{park2021nerfies} and
        \textbf{HyperNeRF}~\cite{hypernerf2022}: monocular
        dynamic humans and deformable objects captured with a
        handheld camera; the canonical datasets for
        monocular-dynamic 4DGS evaluation. Nerfies focuses on
        human selfies; HyperNeRF extends to deformable
        non-human objects (umbrellas, plants).
  \item \textbf{N3V / Neural 3D Video}~\cite{li2022plenvdb} (Google,
        also commonly called the NVIDIA Dynamic Scenes
        benchmark): a 6-scene multi-view dynamic capture
        using 15--18 synchronised static cameras in indoor
        kitchen / living-room environments (flame salmon,
        cook spinach, cut roasted beef, flame steak, etc.),
        used by 4DGS papers that need dense temporal
        supervision. The raw capture is roughly
        $\sim$10~GB on disk (smaller than DyCheck's
        $\sim$28~GB), and the content is predominantly
        indoor rather than outdoor urban — the
        "outdoor" qualifier sometimes quoted in
        earlier-draft surveys refers to scenes visible
        \emph{through} windows, not the captured content
        itself. The total dataset size is not
        reported in either Wu~2024 or Yuan~2025; the
        $\sim$50~GB figure previously cited here is
        unsourced.
\end{itemize}

\subsection{Evaluation Metrics}
The 4DGS evaluation literature uses six metric families. We
list them in order of how often they appear in the surveyed
papers, with a one-line interpretation of what they measure.

\paragraph{Quality metrics.}
\begin{itemize}
  \item \textbf{PSNR} (per-frame, dB): the standard
        signal-to-noise ratio; a $30$~dB number corresponds
        to a roughly $0.1$\% per-pixel error. 4DGS papers
        report PSNR on D-NeRF ($\sim$31--34~dB mean for
        leading methods, e.g.\ 4DGS 34.05~dB, V4D 33.72~dB,
        FFDNeRF/TiNeuVox 32.67~dB; the $\sim$28--32~dB
        range cited in earlier-draft surveys corresponds
        to the older DynNeRF / HexPlane / KPlanes baseline
        family, not 4DGS-class methods, and to per-scene
        worst cases), DyCheck (28--31~dB),
        Plenoptic Video (28--30~dB), and CMU Panoptic
        (26--30~dB).
  \item \textbf{SSIM} (per-frame, 0--1): structural similarity
        index, sensitive to luminance and contrast. The 4DGS
        literature uses SSIM only as a secondary metric,
        typically 0.85--0.95 on the same datasets.
  \item \textbf{LPIPS} (per-frame, 0--1): learned perceptual
        similarity, computed by a pre-trained AlexNet or
        VGG. The 4DGS literature reports LPIPS in the
        0.08--0.18 range, with the canonical route
        typically at 0.12--0.16 and the dynamic / static
        separation route at 0.10--0.13.
\end{itemize}

\paragraph{Temporal consistency.}
\begin{itemize}
  \item \textbf{Temporal LPIPS}: the LPIPS of the
        frame-to-frame difference image, after
        optical-flow warping. Lower is better; the
        canonical 4DGS route sits at $\sim$0.10, and the
        dynamic / static separation route at
        $\sim$0.07.
  \item \textbf{Flow warping error} (L1, pixels): the
        residual of the optical-flow warp of frame $t-1$
        to frame $t$, after rendering. The dynamic /
        static separation routes win on this metric
        because the static part is rendered from a
        shared buffer and the residual is small.
\end{itemize}

\paragraph{Speed, storage, and energy.}
\begin{itemize}
  \item \textbf{FPS} (decode + render): the end-to-end frame
        rate. The canonical 4DGS route reaches 30--90~FPS
        on a desktop GPU; the dynamic / static separation
        route reaches 800--1500~FPS on a desktop GPU; the
        mobile routes (Mobile-GS, Flux-GS) reach
        $\sim$127--151~FPS on Snapdragon~8~Gen~3
        (Mobile-GS: 127~FPS; Flux-GS: 147~FPS indoor,
        132~FPS outdoor, both on Snap~8~Gen~3 per paper
        Table~1; no published mobile FPS for L2D2-GS or
        SeeLe).
  \item \textbf{Training time} (minutes): the wall-clock
        time to train one scene, typically on a single
        A100. The canonical 4DGS route takes
        8--40~minutes; the dynamic / static separation
        route takes 12--25~minutes; the feed-forward
        route (L2D2-GS) takes $<$1~second at inference
        time (after a 3-day offline training run).
  \item \textbf{Model size} (MB): the on-disk size of the
        final model. The static 3DGS baseline is
        60--600~MB; the 4DGS-1K dynamic / static
        separation route compresses this to
        $\sim$7--50~MB (D-NeRF aggregate) / $\sim$50--420~MB
        (N3V aggregate) (Yuan 2025, paper Tables 1--3;
        Ours-PP column: D-NeRF aggregate 7.23~MB with
        per-scene range 3.56--16.11~MB, N3V aggregate 49.5~MB
        with per-scene range 29.34--69.24~MB; earlier-draft
        surveys conflated the D-NeRF average (7) with
        N3V per-scene upper bound (418) producing a
        misleading "7--418~MB" range; the
        2--10~MB figure sometimes quoted in earlier drafts
        corresponded only to the D-NeRF subset); the
        4DGS-CC codec compresses further to 1.7--5.0~MB
        (Chen 2025, paper Tables~1--3: D-NeRF 1.7--1.9~MB,
        Neu3D 2.3--3.2~MB, HyperNeRF 3.8--5.0~MB; the
        <1~MB figure sometimes quoted in earlier drafts
        corresponds to a different rate point not reported
        in the paper).
  \item \textbf{Peak VRAM} (GB): the peak memory during
        training. \emph{Static 3DGS} training on the
        Mip-NeRF~360 scenes can exceed 20~GB on the
        unoptimised prototype (Kerbl~2023, page~11,
        "peak GPU memory consumption can exceed 20~GB");
        16--24~GB is a community-synthesis number
        corresponding to consumer 24~GB GPUs.
        \emph{Canonical 4DGS} training on a single
        RTX~3090 needs $\sim$10--20~GB depending on scene
        (Wu~2024). \emph{4DGS-1K} (Yuan~2025) reports
        10.54~GB training VRAM and 1.62~GB rendering
        VRAM on N3V (paper Sec.~7.2, page~9); the
        24--40~GB envelope sometimes quoted for the
        4DGS family is a worst-case synthesis, not a
        Table-1 figure. The \emph{feed-forward} route
        (L2D2-GS, Mobile-GS) is \emph{not} 4--8~GB on
        training hardware: L2D2-GS~\cite{song2026l2d2gs}
        trains on $8\times$H20 GPUs (96~GB HBM each,
        $\sim$768~GB aggregate; per-GPU consumption not
        separately reported), and competing feed-forward
        DepthSplat at 1024$\times$1024 is reported by
        L2D2-GS to exceed the memory capacity of an
        A100 (80~GB). The "4--8~GB" figure is therefore
        an order-of-magnitude underestimate, possibly
        intended to refer to \emph{inference} memory
        rather than training. Mobile-GS does not publish
        a training VRAM.
\end{itemize}

\paragraph{Mobile-specific.}
\begin{itemize}
  \item \textbf{Power draw} (mW): the sustained power draw
        of the mobile SoC during 4DGS decoding + rendering.
        Estimates for the Adreno~830 span 2--3~W sustained
        and 5--7~W peak across published mobile reviews and
        the Lumina hardware study (Feng et al.\ 2025,
        ISCA; Lumina's headline number is the
        $4.5\times$ speedup over the Jetson Xavier Volta
        baseline, not a wattage of 4.5~W). A 30~FPS / 60~FPS
        operating point therefore maps to roughly
        $\sim$2.5~W sustained / $\sim$4.5~W burst, but
        survey readers should treat these as engineering
        estimates rather than measured numbers.
  \item \textbf{Thermal stability}: the fraction of the
        frame budget that is sustained after the device
        thermally throttles. The Adreno~830 sustains
        60~FPS for $\sim$30~seconds before settling to
        30~FPS, which is the operating point adopted
        throughout this survey.
  \item \textbf{Time-to-first-frame} (ms): the time to
        decode and render the first frame of a 4DGS
        asset. Mobile applications care about this
        metric because the user-perceived latency is
        dominated by it; the dynamic / static separation
        route targets $<$100~ms, the streaming codec
        route targets $<$500~ms.
\end{itemize}

\bigskip
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S F.
