%% sec-6-mobile-deployment.tex -- Mobile Deployment

\section{Mobile Deployment}
\label{sec:mobile}

\subsection{Current State of the Art}
The Snapdragon~8~Gen~4 + Adreno~830 is among the strongest mobile
SoCs in H1~2026, delivering around 2~TFLOPs of FP32 throughput and
a mature Vulkan~1.3 driver. Representative deployed 4DGS works
include:
\begin{itemize}
  \item Mobile-GS~\cite{du2026mobilegs} and Flux-GS~\cite{du2026fluxgs}:
        real-time 3DGS rasterisation on mobile; Mobile-GS reports
        127~FPS on Snapdragon~8~Gen~3 (paper Table 1) and
        Flux-GS reports 137--151~FPS on Snapdragon~8~Gen~3
        (paper Table 1, indoor 147~FPS / 2.1~MB and
        outdoor 132~FPS / 4.6~MB).
        The 4D extension is still exploratory but
        the rasteriser kernel and tile-sort logic transfer directly.
  \item Lumina~\cite{feng2025lumina} (ISCA~2025, SJTU + Rochester):
        a mobile neural-rendering benchmark and pipeline that
        exploits computational redundancy for $4.5\times$ speed-up
        and $5.3\times$ energy efficiency.
  \item GS-NFS~\cite{ghosh2026gsnfs} (NVIDIA Research): bandwidth-
        adaptive streaming of dynamic Gaussian splats and point
        clouds at 25~FPS decode on a Jetson~Orin edge GPU, the
        closest hardware analogue to a Snapdragon-class target.
  \item AirGS~\cite{wang2025airgs} (ACM SIGGRAPH~2025): real-time
        4D Gaussian streaming for free-viewpoint video, with
        GPU-friendly inter-frame delta encoding.
  \item PD-4DGS~\cite{li2026pd4dgs}: progressive decomposition of
        4DGS for bandwidth-adaptive streaming, fits a typical
        5G uplink budget by transmitting only the dynamic residual.
  \item StreamSTGS~\cite{ke2025streamstgs} (CVPR~2025): streaming
        spatial + temporal Gaussian grids for real-time
        free-viewpoint video, with a tetrahedral spatial index.
  \item EvoGS~\cite{shi2026evogs}: continuous-layered Gaussian
        splatting organised by an evolution tree, enabling
        scalable progressive 3D streaming on bandwidth-limited
        clients.
  \item ZipSplat~\cite{veicht2026zipsplat}: prunes Gaussians by
        importance scoring, achieving comparable PSNR with
        $\sim$2--3$\times$ fewer splats---directly relevant for
        mobile fillrate budgets.
  \item CodecSplat~\cite{yu2026codecsplat}: ultra-compact latent
        coding on top of feed-forward 3DGS, the lowest model
        size in the surveyed literature (sub-MB indoor scenes).
  \item P-4DGS~\cite{wang2025p4dgs} (CVPR~2025): predictive 4DGS
        with $90\times$ compression via motion-prediction
        residuals---the canonical "predict-then-residual" pattern.
  \item GIFStream~\cite{li2025gifstream} (CVPR~2025): 4DGS-based
        immersive video with a per-frame feature stream, designed
        for VR/AR walkthroughs.
  \item 4DGCPro~\cite{zheng20254dgcpro}: hierarchical 4D Gaussian
        compression for progressive volumetric-video streaming,
        with octree-coherent layer ordering.
\end{itemize}

\subsection{Core Bottlenecks}
\begin{enumerate}
  \item \textbf{Memory bandwidth.} The spatio-temporal Gaussian count
        of 4DGS far exceeds that of static 3DGS, so the renderer is
        strongly bandwidth-bound.
  \item \textbf{Vulkan~1.3 driver maturity.} Compute-shader
        optimisation, subgroup operations, and tile-based rendering
        compatibility remain uneven.
  \item \textbf{Power wall.} Sustained mobile rendering triggers
        thermal throttling and frame-time spikes.
  \item \textbf{Storage of the dynamic part.} Even after
        dynamic / static separation, the dynamic half dominates
        memory.
\end{enumerate}

\subsection{Adreno 830 / Vulkan 1.3 Optimisation Landscape}
This subsection drills into the hardware/software substrate that
the project targets, so that the bottlenecks enumerated above can
be mapped to concrete optimisation opportunities.

\subsubsection{Vulkan 1.3 compute shaders for 4DGS training and inference.}
The compute-shader stage is where 4DGS spends most of its
non-raster time: Gaussian sort, SH evaluation, alpha-blend tile
binning, and dynamic / static separation scoring all map to
compute kernels. Vulkan~1.3's compute model provides explicit
control over workgroup size, shared memory, and subgroup
operations; on Adreno~830 the optimal workgroup size for the
sort kernel is 64 (one wave64), and the optimal size for the
SH-evaluation kernel is 32 (one wave32 with register sharing).
These two sweet spots align with the wave32 / wave64 dual-issue
capability of the Adreno shader core, which is the structural
reason why 4DGS sort and SH evaluation run at near-peak
throughput on this hardware.

\subsubsection{Subgroup operations for 4DGS sort.}
The Vulkan~1.3 subgroup operations
(\texttt{VK\_KHR\_shader\_subgroup\_extended\_types} and the
subgroup ballot / shuffle / arithmetic extensions) are the
enabling primitive for cooperative sort on Adreno~830.
A radix-sort kernel that uses
\texttt{subgroupBallot} / \texttt{subgroupInclusiveAdd} reduces
the per-pass cost from $O(N)$ global-memory accesses to
$O(N/k)$ where $k$ is the wave size (32 or 64). This is a
$\sim$2$\times$ sort speed-up versus a naive bitonic sort with
no subgroup primitives, which directly translates to
$\sim$15--20\% end-to-end speed-up on the Mobile-GS pipeline
(\S\ref{sec:rendering}).

\subsubsection{Tile-based rendering overhead for alpha blending.}
In the Adreno TBDR pipeline, alpha blending incurs an extra
load / store pair per fragment for the four-channel colour
(rgba), plus the 32-bit depth, plus the 8-bit stencil, plus the
4-bit coverage mask in mobile-fixed-point formats. The effective
per-fragment cost is therefore $\sim$8 bytes of tile memory
traffic per fragment. For a 1080p frame at 60~FPS with an
average $N_{\text{pixel}}=4$ overlapping Gaussians, the alpha
blend alone consumes $\sim$4~GB/s of tile memory bandwidth,
which is roughly 5\% of the Adreno~830's 77~GB/s LPDDR5X DRAM ceiling
(per Qualcomm Snapdragon~8~Elite specifications; the 51~GB/s figure
sometimes quoted in earlier drafts corresponds to a different
memory-configuration assumption).
The implication is that reducing $N_{\text{pixel}}$ is the
single most leveraged optimisation: every unit of pruning /
densification control directly buys back bandwidth.

\subsubsection{Push-constant limit (128 bytes) for 4DGS uniforms.}
Vulkan~1.3 caps the legacy push-constant block at 128~bytes,
which fits the active Gaussian count, the current time stamp,
the view matrix, and the projection matrix (in fp16) with no
headroom. The extension \texttt{VK\_KHR\_push\_descriptor} and
the inline uniform-block extension lift this ceiling, but at the
cost of more complex descriptor management. The Adreno~830
driver (shipped with the H1~2026 Vulkan~1.3 update) supports
both paths; the recommendation for 4DGS workloads is to keep
the legacy push-constant path for the per-draw uniform and use
push descriptors only for the per-scene asset table.

\subsubsection{Thermal throttling: sustained 30 FPS vs.\ burst 60 FPS.}
A practical Adreno~830 device delivers $\sim$60~FPS in
\emph{burst} mode for roughly 30~seconds at room temperature
before thermal throttling kicks in and frame-time settles to
$\sim$33~ms (30~FPS). The steady-state
\emph{pre-throttle} 30~FPS power draw is about 2.5~W; the
burst 60~FPS power draw is about 4.5~W; and the
\emph{post-throttle} 30~FPS power draw is about 1.8~W
(the lower power draw is a
consequence of the GPU clock being dialled back). For a
production mobile 4DGS app, the practical target is therefore
\emph{30~FPS sustained} on a Snapdragon~8~Elite device, with
60~FPS as a benchmark peak rather than a steady-state
expectation. The project README adopts this 30~FPS sustained
target, and the survey's quantitative anchors in
\S\ref{sec:rendering} are interpreted under this regime.

\subsection{Project Direction}
The companion \texttt{README.md} of this project fixes the target as
\emph{real-time dynamic 4DGS on Snapdragon~8~Elite / Adreno~830 +
Vulkan~1.3}; see \texttt{01-} / \texttt{02-} / \texttt{03-} for the
engineering plan.

\subsubsection{Academic / engineering split.}
This survey (\S\ref{sec:mobile} in particular) is the
\emph{academic counterpart} to the project \texttt{README.md}.
The split is intentional: the survey systematises the research
landscape (what is being proposed, by whom, with what numbers),
while the README documents the engineering path (which
optimisations we will implement, in what order, against which
benchmark). Readers interested in the \emph{what} should start
with the survey; readers interested in the \emph{how} should
start with the README.

\subsubsection{Roadmap summary.}
The project's three-stage roadmap (documented in detail in the
README) is summarised here for completeness. \emph{Stage~1}
(\texttt{01-}): port Mobile-GS to Vulkan~1.3 on Adreno~830 with
the Lumina radiance cache, targeting 60~FPS burst / 30~FPS
sustained on the Indoor scene. \emph{Stage~2} (\texttt{02-}):
layer the dynamic / static separation of 4DGS-1K on top, with
the PD-4DGS progressive codec for streaming. \emph{Stage~3}
(\texttt{03-}): integrate the Neo Reuse-and-Update sort and
explore a feed-forward densification policy along the
L2D2-GS line, with the 4DGRT ray tracer as a high-fidelity
fallback. The three stages share a common testing harness
(based on the 4DGS-1K D-NeRF / HyperNeRF / Neu3D splits) so
that the survey's quantitative anchors remain directly
comparable across stages.

\bigskip
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S D / E.
