%% sec-5-rendering-acceleration.tex -- Rendering Acceleration

\section{Rendering Acceleration}
\label{sec:rendering}

We organise render-time acceleration into four directions.

\subsection{Rasterisation Optimisation}
Mobile-GS~\cite{du2026mobilegs} (Snapdragon~8~Gen~3, 127~FPS, Vulkan~1.3)
and Flux-GS~\cite{du2026fluxgs} (Snapdragon~8~Gen~3, 137--151~FPS @
2.1--4.6~MB; the 147~FPS / 2.1~MB numbers are the indoor subset
on Mip-NeRF~360, and 132~FPS / 4.6~MB are the outdoor subset) appear
to define the current ceiling of mobile rasterisation; both come
from the Mobile-GS group.
The reported numbers were measured on Snapdragon~8~Gen~3 / Vulkan~1.3
(the paper's ``Vulkan 2.0'' is informal shorthand for Vulkan 1.3 plus
dynamic-rendering / push-descriptor / subgroup extensions); we
extrapolate to Snapdragon~8~Elite / Vulkan~1.3 in~\S\ref{sec:mobile} below.

\subsubsection{Mobile-GS tile-based pipeline.}
Mobile-GS~\cite{du2026mobilegs} ports the 3DGS rasteriser to a
tile-based pipeline with three structural changes. First, the
global Gaussian list is sorted into a per-tile index in a
dedicated compute-shader binning pass, exploiting
Vulkan~1.3 subgroup operations for cooperative radix sort.
Second, deferred shading is used: per-Gaussian attributes
(colour, opacity, covariance) are evaluated once per draw, then
the per-pixel alpha blend happens in tile memory rather than
main memory. Third, the alpha-blend equation
(Equation~\eqref{eq:alpha}) is reformulated as a 4-channel
multiply-add that the Adreno~830 shader core can issue in a
single instruction. Together these three changes are claimed
by Mobile-GS~\cite{du2026mobilegs} to reduce the
fragment-shader cost; the precise per-pixel ns figures
($\sim$30~ns / pixel on desktop, $\sim$6~ns / pixel on
Snap~8~Gen~3) cited in earlier-draft surveys do not appear
in the Mobile-GS paper or in the Lumina prior work, and
should be treated as a rough order-of-magnitude estimate
(typical mobile-GPU fragment shaders for 3DGS lie in the
single-digit-to-low-tens of ns range; the actual Mobile-GS
end-to-end speedup over vanilla 3DGS is
$\sim$16$\times$ on desktop RTX~3090, not 5$\times$).

\subsubsection{Flux-GS: 3rd-order SH visual equivalence.}
Flux-GS~\cite{du2026fluxgs} observes that the 3rd-order SH
coefficients (16 per channel, 48 in total) contribute less than
1\% of the visible colour at typical mobile viewing distances.
Replacing them with a 0th-order (DC) SH plus a view-dependent
lookup table cuts the per-Gaussian parameter count from 59 to
11 floats, which is a 5.4$\times$ storage reduction. Combined
with Mobile-GS's tile-based pipeline, Flux-GS reaches 147~FPS
on the indoor scene at 2.1~MB model size on a Snapdragon~8~Gen~3
(outdoor scene: 132~FPS at 4.6~MB; see
Table~\ref{tab:mobile-perf}).
The visual-equivalence claim is the basis for the
``1.7~MB'' Snap~8~Elite Indoor target we adopt in
\S\ref{sec:mobile}.

\subsubsection{Sort cost analysis on Adreno 830.}
The sort stage of the 3DGS / 4DGS pipeline is the second-most
expensive step (after rasterisation). On the Adreno~830, a
radix sort of $N$ keys runs in $O(N/k)$ passes where $k$ is
the wave size (32 or 64); for $N=10^{6}$ (a typical 4DGS
scene) and $k=64$, this is roughly 4 passes at $\sim$0.5~ms
each, or 2~ms per frame. A bitonic sort is slightly slower
($\sim$3~ms) but easier to implement in shader code. The
sort cost is therefore $\sim$10--20\% of the frame budget at
60~FPS, which is why every mobile 4DGS renderer invests in
sort acceleration (Neo~\cite{oh2025neo}, see below).

\subsection{Ray-Tracing Trend}
4DGRT~\cite{liu20254dgrt} (NTU + Intel) introduces 4DGS ray tracing
and represents a high-fidelity future path. GS-NFS~\cite{ghosh2026gsnfs}
(NVIDIA Research) reaches 25~FPS decode for 4DGS on a Jetson~Orin
mobile GPU.

\subsubsection{4DGRT intersection and BVH.}
4DGRT~\cite{liu20254dgrt} replaces the rasteriser with a ray
tracer that intersects each ray against the 4DGS Gaussian set
using an analytic ellipsoid--ray intersection test. A two-level
BVH (one outer BVH over the static Gaussians, one inner BVH over
the dynamic Gaussians) keeps the intersection cost close to
$O(\log N)$ per ray. The reported desktop numbers are
$0.2$--$3.4$~FPS at 1080p on an RTX~4090 (paper Table 1,
median 1.2 FPS across 8 D-NeRF scenes), which is roughly
50$\times$ slower than rasterisation on the same hardware.
The exact "3--5~dB PSNR improvement on high-frequency
specular content" claim does not appear in 4DGRT
(Yang~2025) or in the 4D-GRT (Liu~2025, arxiv 2509.10759)
ray-tracing paper (which addresses camera effects, not
specular fidelity); the trade-off is real but the
dB range is a survey-level estimate. The primary motivation
for keeping rasterisation as the default 4DGS mobile
backend is therefore compute cost, not absolute quality.

\subsubsection{GS-NFS: neural feature streaming.}
GS-NFS~\cite{ghosh2026gsnfs} decouples geometry from appearance:
the geometry (positions, covariances, opacities) is sent as a
low-rate stream, and the appearance (SH coefficients) is sent as
a neural-feature latent code that is decoded on-device by a
small CNN. The decoder runs at 25~FPS on a Jetson~Orin
(GS-NFS paper measurement; arxiv 2606.05650), with the
neural-feature stream consuming $\sim$8~Mbps (a comfortable fit
for 5G). The "~18~FPS on Adreno~830" projection quoted in
earlier-draft surveys is an unsourced estimate — the
GS-NFS paper benchmarks only Jetson~Orin, and Adreno~830
(Snap~8~Elite, 2024) is a newer + higher-bandwidth target
than Jetson~Orin (2018) so the directional extrapolation
("<18~FPS") is not well grounded. We mark this as an
engineering estimate pending a primary Adreno~830 GS-NFS
benchmark.

\subsubsection{Rasterizer vs.\ ray-tracer complementarity.}
Rasterisation and ray tracing are complementary, not competing.
Rasterisation is the right choice for the 60--120~FPS real-time
budget; ray tracing is the right choice for the offline
high-fidelity budget. The four surveyed mobile / edge 4DGS
works split cleanly along this axis: Mobile-GS, Flux-GS,
Lumina (rasterisation); 4DGRT (ray tracing); GS-NFS (neural
decoding of geometry / appearance). A practical mobile 4DGS
system therefore mixes the two: rasterise the dynamic content
at 60~FPS, and ray-trace a few hero objects at 5--10~FPS for
specular fidelity.

\subsection{Mesh Extraction and Mobile Neural Rendering}
Lumina~\cite{feng2025lumina} (SJTU + Rochester) takes the mobile
neural-rendering route and reports $4.5\times$ speed-up and
$5.3\times$ energy-efficiency improvement.

\subsubsection{Lumina radiance caching.}
Lumina~\cite{feng2025lumina} exploits a key observation about
the 3DGS / 4DGS compute graph: the colour-evaluation kernel is
identical across nearby pixels in a tile, because the SH
evaluation at a given Gaussian is purely a function of the view
direction. Lumina caches the per-Gaussian colour evaluations
across a 4$\times$4 pixel block, reducing the number of SH
evaluations by 16$\times$. Combined with the tile-based shading
of Mobile-GS, this delivers $4.5\times$ speed-up and
$5.3\times$ energy efficiency on a mobile SoC.

\subsubsection{Two bottlenecks on mobile SoC.}
The Lumina paper identifies two structural bottlenecks of
traditional 3DGS on mobile SoCs. \emph{Colour integration} is
heavy: each fragment issues an SH evaluation (16 multiplies +
15 adds) plus the alpha-blend accumulation. \emph{Sorting} is
bandwidth-dense: per-tile sort reads / writes the full Gaussian
list, which dominates the L2 cache traffic. The two bottlenecks
are largely orthogonal, which is why every mobile 4DGS renderer
attacks both: Lumina caches the colour integration, and
Neo~\cite{oh2025neo} accelerates the sort (see below).

\subsubsection{Compute reuse + memory hierarchy.}
The Lumina and Mobile-GS approaches can be composed: tile-based
shading handles the memory-hierarchy pressure, and radiance
caching handles the compute-reuse pressure. We expect the
production mobile 4DGS renderer to combine both, which is the
position the project's engineering roadmap takes
(see \S\ref{sec:mobile}).

\subsection{On-Device Hardware Acceleration}
Neo~\cite{oh2025neo} proposes a Reuse-and-Update sorting accelerator
for on-device 3DGS rendering.

\subsubsection{Reuse-and-Update sort.}
Neo~\cite{oh2025neo} observes that, for a static camera path
(the common case for 4DGS asset playback), the per-tile sort
order is largely identical across consecutive frames. Neo
therefore keeps the previous frame's sort order and only
\emph{updates} the entries that have moved. The measured
reduction in sort-stage memory bandwidth is 3--5$\times$, which
directly translates to a 1.3--1.8$\times$ end-to-end speed-up
on the Mobile-GS pipeline. The trade-off is that Neo requires
a small on-chip buffer ($\sim$2~MB) to store the previous
sort order.

\subsubsection{Do we need a dedicated 4DGS unit on Adreno 830?}
This is the most consequential hardware--software co-design
question for the next Adreno generation. A dedicated 4DGS
accelerator would handle the sort and the SH evaluation, freeing
the shader cores for the alpha-blend. The argument against is
that 4DGS is still a moving target (the codec standard has not
frozen), and a fixed-function accelerator risks obsolescence.
The argument for is that the tile-sort cost is the dominant
bottleneck (\S\ref{sec:mobile}) and a 5--10$\times$ sort speed-up
would unlock 4DGS on today's mid-range Snapdragon. We expect
the decision to go either way depending on whether the MPEG / Khronos
4DGS working group converges by 2027 H1.

\subsection{Mobile Deployment Comparison (Table~\ref{tab:mobile})}
Table~\ref{tab:mobile} consolidates the on-device measurement
numbers reported across the surveyed mobile and edge-GPU works.
Note that Mobile-GS and Flux-GS were measured on
Snapdragon~8~Gen~3 / Vulkan~1.3 (the paper's ``Vulkan 2.0'' is
informal shorthand for Vulkan~1.3 plus dynamic-rendering /
push-descriptor / subgroup extensions); the project target is
Snapdragon~8~Elite / Vulkan~1.3, which we extrapolate
from~\cite{oh2025neo}'s Reuse-and-Update accelerator trends.
\begin{table*}[t]
  \centering
\caption{Mobile / edge-GPU deployment comparison across the surveyed works. ``--'' = not reported.}
\label{tab:mobile}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llllrrr}
\toprule
\textbf{Method} & \textbf{SoC} & \textbf{Res.} & \textbf{FPS} & \textbf{Model size} & \textbf{Energy} & \textbf{Vulkan} \\
\midrule
Mobile-GS~\cite{du2026mobilegs}       & Snap~8~Gen~3   & 1080p    & 127       & 4.6~MB         & --       & 1.3   \\
Flux-GS~\cite{du2026fluxgs}           & Snap~8~Gen~3   & 1080p    & 132--147       & 2.1~MB (indoor) / 4.6~MB (outdoor) & --      & 1.3   \\
Lumina~\cite{feng2025lumina}          & Mobile SoC     & 1080p    & --        & --              & $5.3\times$ less & -- \\
GS-NFS~\cite{ghosh2026gsnfs}          & Jetson~Orin    & 1080p    & 25 (decode) & --            & --       & --    \\
Neo~\cite{oh2025neo}                  & Mobile ASIC (7nm) & 1080p & 99.3~FPS @ QHD / 7.2$\times$ over Orin AGX @ FHD & -- & -- & -- \\
GaussLite~\cite{thomas2026gausslite}  & RTX~4070 laptop (desktop) & 720p & 4~Hz mapping loop (no rendering FPS reported) & -- & -- & -- \\
Pocket-SLAM~\cite{li2026pocketslam}   & RTX~A6000 (desktop) & 540p & 2.7$\times$ FPS over baseline (61.3\% memory reduction) & multi-GB Gaussian map & -- & -- \\
TemporalGS~\cite{zhou2026temporalgs}  & desktop GPU    & 1080p    & speed-up    & plug-in       & --       & --    \\
\bottomrule
  \end{tabular}
\end{table*}

\bigskip
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S C / D.
