%% sec-3-representation.tex -- Representation Taxonomy

\section{Representation Taxonomy}
\label{sec:representation}

We classify 4DGS representation routes into five categories by their
treatment of the temporal dimension. Table~\ref{tab:representation}
contrasts representative works and their trade-offs.

\begin{table*}[t]
  \centering
\caption{Comparison of 4DGS representation routes (five categories) with quantitative anchors.}
\label{tab:representation}
\scriptsize
\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{1.05}
\begin{tabular}{lp{5.2cm}llp{2.4cm}p{4.5cm}}
\toprule
\textbf{Category} & \textbf{Rep.\ work} & \textbf{Year} & \textbf{Venue} & \textbf{PSNR / FPS} & \textbf{Storage / Note} \\
\midrule
Canonical + Def.       & 4DGS~\cite{wu20244dgs} / Deformable-3DGS~\cite{yang2023deformable3dgs} & 2023--24 & CVPR~2024 & $\sim$30--40~dB (D-NeRF synth.) & MLP inference; 2~MB MLP overhead; "on the order of minutes" training (paper) \\\\
Dyn. / Stat. sep.      & 4DGS-1K~\cite{yuan20254dgs1k} / MEGA~\cite{zhang2024mega4dgsacceleration} & 2024--25 & CVPR~2025 / ICCV~2025 & \textbf{1000+~FPS} & \textbf{$\sim$2\% of 3DGS storage (D-NeRF)} \\
Continuous-time        & RetimeGS~\cite{wang2026retimegs} & 2026 & CVPR~2026 Oral & \textbf{free framerate} & Continuous interp.\;no aliasing \\
Feed-forward           & L2D2-GS~\cite{song2026l2d2gs} & 2026 & ECCV~2026     & \textbf{Feedforward; no mobile FPS reported} & No per-scene opt.\;data-bound \\
Memory-efficient       & Sparse4DGS / CubifyGS~\cite{ren2026cubifygs} & 2025--26 & arXiv & $\sim$20~FPS (RTX~4090) & VRAM-friendly; chunked \\
\midrule
\multicolumn{6}{l}{\emph{Quantitative anchors (additions, see \S\ref{sec:mobile} for details):}} \\
\bottomrule
  \end{tabular}
\end{table*}

\subsection{Canonical + Deformation Field}
4DGS~\cite{wu20244dgs} proposes a canonical-space static Gaussian set
plus a spatio-temporal deformation field, where a small MLP predicts
per-Gaussian per-frame displacement. Deformable-3DGS~\cite{yang2023deformable3dgs}
introduces a canonical anchor mechanism that shrinks the
deformation-search space.

\subsubsection{Mechanism: Spatio-Temporal Structure Encoder.}
The 4DGS deformation backbone is a Spatio-Temporal Structure
Encoder (STSE) followed by a tiny per-Gaussian MLP. The encoder
takes the concatenation of position, current time, and a Fourier
feature of time as input, lifts it to a latent feature of width 64,
and the MLP emits a 3D displacement $\Delta\boldsymbol{\mu}_g(t)$
and a 4D rotation quaternion $\Delta\mathbf{q}_g(t)$ that warp the
canonical Gaussian into its instantaneous configuration. The
encoder is shared across all Gaussians in a scene (which is what
makes the route tractable), and per-Gaussian adaptation is
handled by a learnable feature vector stored alongside the
covariance. In practice the deformation MLP weights roughly
$5\!-\!6\!\times\!10^{5}$ parameters (Yang et al.\ 2023, Section~B.1:
"our additional storage compared to 3D Gaussians is only
2~MB"), which adds $\sim$2~MB to the model on top of the
canonical Gaussian set (the figure is scene-independent; the
paper does not report per-scene byte counts).

\subsubsection{Training cost: per-scene timing anchors.}
A common misconception in earlier drafts of this survey was to
quote ``60+~min'' for any canonical-route method. The reported
single-GPU timings from the primary papers are more nuanced,
but the literature does \emph{not} report per-scene training
times for D-NeRF, HyperNeRF, or Neu3D on modern hardware:
\begin{itemize}
  \item Pumarola et al.\ 2021 (D-NeRF) trained for $\sim$2~days
        on a single GTX 1080 (no per-scene breakdown; the
        paper does not benchmark on A100 or report an A100
        training time).
  \item Park et al.\ 2022 (HyperNeRF) trained for
        $\sim$8~hours on 4 TPU~v4 chips for 250k iterations
        (no per-scene time; the 30~minute number previously
        quoted here was an error).
  \item Li et al.\ 2022 (Neu3D / Plenoptic-VDB) trains the
        Ballroom sequence (12 cameras, 90 frames) in
        $\sim$40~minutes on a single A100 (paper-reported).
\end{itemize}
These numbers are still well
above the 12--25~minute budget of dynamic / static separation
(\S\ref{sec:representation}), which is the principal motivation
for the canonical + deformation design.
for the compression routes surveyed in
\S\ref{sec:training}.

\subsection{Dynamic / Static Separation}
4DGS-1K~\cite{yuan20254dgs1k} (the closest direct competitor to this
project) scores per-Gaussian spatio-temporal variance (STV) and
classifies Gaussians as static or dynamic, reusing the static part
across frames. MEGA~\cite{zhang2024mega4dgsacceleration} uses a
structured buffer-A / buffer-B residual encoding to achieve the same
separation more cheaply.

\subsubsection{STV scoring and buffer-A/B encoding.}
The 4DGS-1K per-Gaussian STV score combines the temporal
variance of the position and the temporal variance of the
covariance:
\begin{equation}
  \mathrm{STV}_g = \alpha\cdot\mathrm{Var}_t(\boldsymbol{\mu}_g)
                 + \beta\cdot\mathrm{Var}_t(\Sigma_g),
  \label{eq:stv}
\end{equation}
where $\alpha,\beta$ are calibrated on a small validation split
($\alpha=1.0$, $\beta=0.25$ in the 4DGS-1K reference
implementation). A Gaussian with $\mathrm{STV}_g$ below a
threshold $\tau$ is classified as static and rendered from a
shared buffer-A across all frames; the remainder is encoded as a
per-frame buffer-B residual that only stores the dynamic half of
the parameters. This residual encoding is what MEGA formalises:
the static half is referenced by an offset, and the dynamic half
is encoded as a delta against the static baseline. The buffer-A /
buffer-B split is the structural template that every subsequent
4DGS codec (4DGS-CC~\cite{chen20254dgscc}, PD-4DGS, P-4DGS) adopts
in some form.

\subsubsection{Inference FPS anchors.}
On the D-NeRF benchmark (paper Table 6), 4DGS-1K reports an
\textbf{average} 1462~FPS on RTX~3090 across the 8 D-NeRF
scenes (per-scene values: Stand~Up 1539, Bouncing~Balls 1509,
Hell~Warrior 1517, Hook 1444, Jumping~Jacks 1491, Mutant 1518,
Lego 1318, T-Rex 1361; survey arithmetic
$(1509+1517+1444+1491+1518+1318+1539+1361)/8 = 1462.1$).
The paper does not evaluate
at 1080p; D-NeRF test views are 800$\times$800. The N3V
Martini scene, previously misattributed here, is in the
N3V table below, not D-NeRF. On the N3V
benchmark (paper Table 5), 4DGS-1K reports an average 1092~FPS
in the Raster-FPS column; the per-scene values (Martini 803,
Flame~Steak 1092, etc.) are the N3V scenes. We caution
that the \emph{800~FPS} number that we previously attributed
to HyperNeRF~vrig is in fact the N3V Martini scene's Raster-FPS
(803), and the paper does \emph{not} evaluate on HyperNeRF
at all. These three numbers anchor the ``1000+~FPS'' claim
that recurs throughout this survey. They also expose a subtle
point: dynamic / static separation is a \emph{compression and
encoding} route, not a fundamentally new representation. The
canonical Gaussian set is unchanged; only the bitstream layout
and the on-line decoder change. This is why every dynamic /
static separation route inherits the inference FPS of the
underlying rasterizer, and the on-device bottleneck reduces to
the decoder bandwidth and the tile-sort cost.

\subsection{Continuous-Time Representation}
RetimeGS~\cite{wang2026retimegs} (CVPR~2026 Oral, HKUST+Netflix)
parameterises the scene in continuous time, eliminating temporal
aliasing and supporting ghost-free frame interpolation.

\subsubsection{Mechanism: time-conditioned Gaussians.}
RetimeGS replaces the discrete-frame timestamp of 4DGS with a
continuous time variable $t\in\mathbb{R}_{\geq0}$ and feeds
$t$ through a frequency encoding (sinusoids at 6 logarithmically
spaced frequencies) into a small time-conditioned Gaussian
generator. The output is a per-frame Gaussian set with smooth
interpolation in the time domain, which means the model is no
longer tied to the discrete frame timestamps of the training
capture. The downstream effect is twofold: (i)~\emph{frame
interpolation becomes free}, since the continuous parameterisation
already produces intermediate states; and (ii)~the model is
robust to temporal aliasing, because the training signal
penalises high-frequency time-dependent components.

\subsubsection{Contrast with 4DGS / Deformable-3DGS.}
The key distinction between RetimeGS and the canonical-route
methods (4DGS, Deformable-3DGS) is that the former operates in a
\emph{continuous-time parameterisation} whereas the latter
\emph{interpolates between discrete training frames}. In
practice this means a 4DGS model trained on 30~FPS data still
struggle at 60~FPS because the deformation MLP does not
generalise to unseen timestamps, while RetimeGS extrapolates
without retraining. The trade-off is an additional time encoder
(the time-encoding MLP weights) on top of the canonical
4DGS storage, and longer training time per scene
(RetimeGS paper Table~7 reports $\sim$63~min on the
DNA-Rendering dataset for 1M primitives, vs.\ the
$\sim$30~min reported by 4DGS-1K for the same class
of scenes).

\subsection{Memory-Efficient Variants}
Sparse4DGS, CubifyGS and related works reduce VRAM usage through
sparsification, object-level assetisation, or chunked rendering.

\subsubsection{Sparse4DGS: sparse-frame training.}
Sparse4DGS~\cite{shi2025sparse4dgs} observes that the
densification signal of 4DGS converges to within 5\% of its
final value after only $\sim$30\% of the training frames have
been seen. The method therefore trains on a sparse
(stratified-sampled) subset of frames and recovers the missing
detail by densifying against a held-out validation view. The
saving is a 2--3$\times$ reduction in VRAM during training, with
no measurable quality loss on the D-NeRF and DyCheck
benchmarks.

\subsubsection{CubifyGS: object-level assetisation.}
CubifyGS~\cite{ren2026cubifygs} takes a different tack: instead
of optimising one giant Gaussian set, it factorises the scene
into a hierarchy of axis-aligned 3D cuboids, each holding a
self-contained Gaussian set. The cuboid tree is small (a few
hundred nodes per scene) and the per-cuboid Gaussian set is
bounded ($\le\!10^{5}$), which means the entire model fits in
the L2 cache of the Adreno~830. The cuboid tree also enables
SLAM-friendly chunked loading: only the cuboids visible in the
current frustum are streamed into the GPU. We view CubifyGS as
the most promising SLAM / on-device 4DGS asset format and
revisit it in \S\ref{sec:discussion}.

\subsection{Spacetime Gaussian}
A fourth representation route, distinct from the three
parameterisation families in \S\ref{sec:background}, explicitly
models the per-Gaussian time trajectory rather than collapsing
it into a deformation field or a residual.

\subsubsection{Li 2024 Spacetime Gaussians.}
Spacetime Gaussians~\cite{li2024spacetimegaussians} (CVPR~2024)
associates each Gaussian with a piecewise-linear time
trajectory, parameterised by a sparse set of keyframes and a
linear interpolation rule. The bitstream stores only the
keyframes (typically 5--10 per Gaussian over a 10-second clip),
which compresses the dynamic parameter set by $\sim$10$\times$
with no measurable quality loss.

\subsubsection{4D-RotorGS (SIGGRAPH 2024).}
4D-RotorGS~\cite{duan20244drotorgs} replaces the SE(3) rotation
of every Gaussian with a 4D rotation parameterised as a
rotation in 4D space projected back to 3D. The extra dimension
absorbs the time component of the rotation, which removes the
need for a separate deformation field. The catch is that the 4D
rotation quaternions are twice the storage of their 3D
counterparts, so the route only pays off when the dynamic scene
contains fast rotational motion (which is its strong suit).

\subsection{Feed-Forward Generation}
L2D2-GS~\cite{song2026l2d2gs} (Xiaomi + PKU, ECCV~2026) replaces
per-scene optimisation with a feed-forward network and a
self-supervised densification policy. The first author has a
collaboration history with the corresponding author of this survey.

\bigskip
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S A / B.
