%% sec-2-background.tex -- Background

\section{Background}
\label{sec:background}

\subsection{3D Gaussian Splatting}
3DGS~\cite{kerbl20233dgs} represents a scene as a set of anisotropic
3D Gaussian ellipsoids. Each Gaussian is parameterized by its
position, covariance, opacity, and spherical-harmonic coefficients.
The renderer projects Gaussians to screen space via a differentiable
rasterizer and composites per-pixel color using alpha blending.

\paragraph{Representation.}
A single 3D Gaussian is described by a mean position
$\boldsymbol{\mu} \in \mathbb{R}^{3}$, a $3 \times 3$ covariance
matrix $\Sigma$ (factored as $\Sigma = R S S^{\mathsf{T}} R^{\mathsf{T}}$
with rotation $R$ and scale $S$ to keep the matrix positive
semi-definite), an opacity $\alpha \in [0, 1]$, and a band-limited
spherical-harmonic (SH) coefficient tensor that encodes
view-dependent colour. The full parameter count per Gaussian is
$3 + 7 + 1 + 3 k^{2}$ where $k$ is the SH band (typically
$k = 0$ for static scenes, $k = 3$ for view-dependent rendering).
At $k = 3$ each Gaussian occupies roughly $59$~bytes, and a typical
urban-scale scene is reconstructed with $1$--$10$~million
Gaussians, yielding a model size of $60$--$600$~MB. Recent
work~\cite{yu2024mipsplatting, fan2023lightgaussian, chen2024hacpp}
shows that the SH component dominates (about $78\%$ of the bytes)
and is the primary compression target.

\paragraph{Rendering.}
Rendering a 3D Gaussian onto the image plane is a two-step
operation. First, the 3D covariance is projected to a 2D screen-space
covariance $\Sigma'$ via the EWA splatting equation
$\Sigma' = J W \Sigma W^{\mathsf{T}} J^{\mathsf{T}}$, where $J$ is
the Jacobian of the projective transformation and $W$ is the
viewing matrix; this is the same analytic approximation introduced
in Zwicker et al.'s 2001 EWA surface splatting, and it preserves the
elliptical footprint of each Gaussian in image space. Second, the
2D Gaussian is rasterised onto screen tiles and composited with the
back-to-front $\alpha$-blending rule
$\mathbf{C} = \sum_{i} \mathbf{c}_{i} \alpha_{i} \prod_{j < i}
(1 - \alpha_{j})$, which is fully differentiable and admits
gradient-based densification. Compared to NeRF-style volume
rendering, which draws $N = 100$--$800$ samples along each ray
through a 5D position-and-direction MLP, 3DGS evaluates exactly
one $\alpha$-blend per contributing Gaussian; on a desktop GPU
this yields a 50--$100\times$ speed-up at 1080p, which is the
empirical foundation of the entire 4DGS research programme
discussed in this survey.

\subsubsection{Mathematical formulation.}
The 2D screen-space covariance follows the EWA splatting
projection of Zwicker~\emph{et~al.}~\cite{kerbl20233dgs}:
\begin{equation}
  \Sigma' = J\,W\,\Sigma\,W^{\top}J^{\top},
  \label{eq:ewa}
\end{equation}
where $W$ is the view transform and $J$ is the Jacobian of the
perspective projection. Equation~\eqref{eq:ewa} guarantees that
the projected ellipse is still positive semi-definite, which is a
prerequisite for the per-pixel analytic footprint used by the
3DGS differentiable rasterizer. The colour at view direction
$\mathbf{d}$ is then evaluated via SH band-limited reconstruction:
\begin{equation}
  C(\mathbf{d}) = \sum_{\ell=0}^{L}\sum_{m=-\ell}^{\ell}
                 c_{\ell}^{m}\,Y_{\ell}^{m}(\mathbf{d}),
  \label{eq:sh}
\end{equation}
where $Y_{\ell}^{m}$ are the real spherical-harmonic basis
functions. In practice, 3DGS uses $L=3$, giving 16 SH coefficients
per colour channel; together with the 3D position (3 floats), the
7 floats of the quaternion / scale, the opacity (1 float), and the
48 SH floats ($16\times3$ channels) this amounts to roughly 59
floating-point parameters per Gaussian. When bloat from
densification is counted, the practical per-Gaussian footprint
climbs into the 100--200-float range, a number that directly
motivates the compression routes discussed in
\S\ref{sec:training}.

\subsubsection{Volumetric rendering vs.\ splatting.}
Compared to NeRF-style volumetric rendering, which integrates
colour along each ray through $N\!\in\![100,800]$ densely-sampled
points (and is therefore compute-bound by ray-marching depth),
3DGS uses a single per-Gaussian rasterization pass followed by
front-to-back alpha compositing:
\begin{equation}
  \mathbf{C}_{\text{pixel}} = \sum_{i=1}^{N_{\text{pixel}}}
      T_i\,\alpha_i\,\mathbf{c}_i, \qquad
  T_i = \prod_{j=1}^{i-1}\bigl(1-\alpha_j\bigr),
  \label{eq:alpha}
\end{equation}
where $T_i$ is the accumulated transmittance and $N_{\text{pixel}}$
is the number of Gaussians overlapping the pixel. This shifts the
dominant cost from per-ray volume sampling to per-tile sorting and
fragment shading---precisely the cost structure that
\S\ref{sec:rendering} and \S\ref{sec:mobile} attack.

\subsection{4D Gaussian Splatting}
4DGS extends 3DGS with an explicit time dimension. The dominant
parameterization routes fall into three categories:
\begin{itemize}
  \item \textbf{Canonical + deformation field.}
        4DGS~\cite{wu20244dgs} and Deformable-3DGS~\cite{yang2023deformable3dgs}
        keep a canonical static representation and apply a per-frame
        deformation MLP.
  \item \textbf{Dynamic / static separation.}
        4DGS-1K~\cite{yuan20254dgs1k} and MEGA~\cite{zhang2024mega4dgsacceleration}
        score per-Gaussian spatio-temporal variance (STV) or encode
        residuals with a buffer-A/B scheme.
  \item \textbf{Continuous-time representation.}
        RetimeGS~\cite{wang2026retimegs} eliminates temporal aliasing
        and enables ghost-free frame interpolation.
\end{itemize}

\subsubsection{Trade-off anchors across the three routes.}
Table~\ref{tab:bg-tradeoffs} contrasts the three parameterisation
families with the quantitative anchors used throughout the rest of
this survey. The numbers come from the primary papers
(canonical~\cite{wu20244dgs,yang2023deformable3dgs}, dyn./stat.\
separation~\cite{yuan20254dgs1k,zhang2024mega4dgsacceleration},
continuous-time~\cite{wang2026retimegs}) and are revisited in
\S\ref{sec:representation}.

\begin{table*}[t]
  \centering
\caption{Trade-off anchors across the three dominant 4DGS
parameterisation families (canonical + deformation, dynamic /
static separation, continuous-time).}
\label{tab:bg-tradeoffs}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrr}
\toprule
\textbf{Family} & \textbf{Parameters / frame} &
\textbf{Train time} & \textbf{Inference FPS} &
\textbf{Typical storage} \\
\midrule
Canonical + Def. & $\sim$2$\times$ static & 8--40~min/scene & 30--60 & 50--200~MB \\
Dyn./Stat.\ sep. & 1$\times$ (shared) + $\epsilon$ & 12--25~min & 800--1500 & $\sim$2--10~MB \\
Continuous-time & $\sim$1.5$\times$ static + $t$-embed & 20--45~min & free framerate & 30--80~MB \\
\bottomrule
  \end{tabular}
\end{table*}

\subsection{Mobile GPU Foundations}
Mobile GPUs such as the Adreno 830 are constrained by: compute
budget (around 2~TFLOPs FP32), memory bandwidth, the maturity of
Vulkan~1.3 drivers, and the cost of alpha blending on the
tile-based rendering architecture.

\subsubsection{Vulkan 1.3 mobile features.}
Vulkan~1.3 (released 25~January~2022, ratified for Android in
late 2023) is the first Vulkan profile that ships with consistent
support across the Adreno, Mali, and Xclipse mobile GPUs. Three
sub-features matter most for 4DGS. \emph{Subgroup operations}
(\texttt{VK\_KHR\_shader\_subgroup\_extended\_types} and
\texttt{VK\_KHR\_shader\_subgroup\_uniform\_control\_flow}) expose
wave32 / wave64 SIMD lanes; the Adreno~830 shader core supports
both wave sizes, which directly enables the cooperative
bitonic / radix sort used by 4DGS per-tile sorting. \emph{Push
constants} (\texttt{VK\_KHR\_push\_descriptor} and the inline
uniform-block extension) lift the legacy 128-byte push-constant
ceiling for tightly-coupled per-draw uniforms (e.g. the active
Gaussian count and current time stamp), which we discuss further
in \S\ref{sec:mobile}. \emph{Descriptor indexing}
(\texttt{VK\_EXT\_descriptor\_indexing}) allows unbounded bindless
texture / buffer arrays, which is the data structure on which the
mobile 4DGS asset cache is built.

\subsubsection{Adreno 830 architecture and the tile-based pipeline.}
The Adreno~830 (Snapdragon~8~Elite, Qualcomm H1~2026) integrates
six unified shader cores clocked at up to 1.25~GHz
(standard profile; the Samsung-exclusive boost profile reaches
$\sim$1.35~GHz), with a 12~MB shared L2 cache and
an effective DRAM bandwidth of about 77~GB/s over LPDDR5X
(per Qualcomm Snapdragon~8~Elite specifications; the 51~GB/s
figure sometimes quoted in earlier drafts corresponds to a
lower-speed LPDDR5X configuration). Crucially, the Adreno
follows a tile-based deferred-rendering (TBDR) architecture:
the screen is split into
$4\!\times\!4$ bins of $64\!\times\!64$ tiles, Gaussians are
binned by the geometry stage, and per-tile resolve passes composite
fragments into tile memory before the final framebuffer write. In
this pipeline, alpha-blending incurs an extra load/store pair per
fragment (4~channels of colour plus the 32-bit depth, plus the
8-bit stencil and the 4-bit coverage mask in mobile-fixed-point
formats), and the per-pixel blending cost is therefore
proportional to the number of overlapping Gaussians $N_{\text{pixel}}$
from Equation~\eqref{eq:alpha}. Reducing $N_{\text{pixel}}$ is the
single most leveraged optimisation for mobile 4DGS, which is why
pruning (\S\ref{sec:training}), binning (\S\ref{sec:rendering}),
and tile-sort acceleration (\S\ref{sec:rendering}) all appear
prominently below.

\subsubsection{Methodology recap: 5-step scan and 9-block notes.}
The corpus of 50+ works in this survey was assembled through a
five-step pipeline. \emph{Step~1:} a curated keyword / author
whitelist (5G-4DGS, mobile-GS, retime, EvoGS, etc.). \emph{Step~2:}
a weekly \texttt{arxiv\_4dgs\_scan} cron that scans
\texttt{cs.CV} / \texttt{cs.GR}. \emph{Step~3:} a 1-hop citation
walk from the seed papers 3DGS~\cite{kerbl20233dgs} and
4DGS~\cite{wu20244dgs}, plus DBLP and Semantic-Scholar lookups for
citation chains. \emph{Step~4:} each candidate is summarised in a
nine-block paper note (title / venue / faction / one-line
contribution / mechanism diagram / quantitative numbers /
mobile-implication flag / open-source link / 1-hop neighbours).
\emph{Step~5:} a cascade propagation step that re-emits the
updated note into \texttt{survey.bib} and the per-section
\verb|\item| block. The 9-block template, the 1-hop rule, and the
5-step scan are inherited from the project's
\texttt{INDEX.md}~\S A and replicated in every section footer.

\subsection{Methodology of this Survey}
This survey is the result of a two-year literature scan focused on
real-time dynamic 4DGS rendering on mobile GPUs. The corpus was
assembled from four primary sources: the arXiv listings under
\texttt{cs.CV} and \texttt{cs.GR} (queried weekly via the project's
\texttt{arxiv\_4dgs\_scan} cron job, which filters by a curated
whitelist of keywords and known authors), the DBLP and Semantic
Scholar indices cross-checked for citation chains, the CGF / TOG
/ SIGGRAPH / CVPR / ICCV / ECCV / ICLR / ICML / NeurIPS
proceedings between 2023 and 2026, and the public benchmarks and
code repositories (GitHub) of the 50+ primary works discussed
below. Inclusion criteria are: (i)~the work targets dynamic scene
reconstruction or rendering; (ii)~it includes either a quantitative
mobile or on-device evaluation, or a clearly-articulated design
implication for mobile deployment; (iii)~a reproducible artifact
(code, paper-note, or detailed PDF) is available. Exclusion
criteria are: pure theoretical contributions without an
implementation, works published in languages other than English,
and works superseded by a more recent version of the same
contribution (we cite the latest). The 50+ primary works are
organised into a five-faction taxonomy (see Figure~\ref{fig:taxonomy})
and three pipeline stages; the resulting 9-section structure is
described in §1. The survey is regenerated on every push to
\texttt{main} via a GitHub Action that compiles the PDF and
publishes it to the project GitHub Pages; the underlying
\texttt{scripts/assign\_survey\_section.py} and
\texttt{scripts/extract\_paper\_summary.py} tools keep the
\texttt{survey\_section} field, the per-section \verb|\item|
blocks, and the \texttt{survey.bib} bibliography in sync with the
54 paper notes.

\subsection{Related Surveys and Positioning}
Three prior surveys overlap with this work; we position our
contribution relative to each. \emph{Fei et al. (2024) ``3DGS:
Survey, Technologies, Challenges, and Opportunities''}
(arXiv 2407.17418) covers 3DGS broadly but does not address
dynamic scenes or mobile deployment. \emph{Chen \& Wang (2024)
``Recent Advances in 3D Gaussian Splatting''} (arXiv 2403.11134)
focuses on 3D reconstruction, editing, and downstream
applications; it also predates most of the 4DGS literature.
\emph{Dellaert et al. (2024) ``A Survey on 3D Gaussian
Splatting''} (arXiv 2401.03820) is a foundational
terminology survey with limited depth. The present survey
differs in three ways: (1)~it is the first to focus
exclusively on \emph{dynamic} 4DGS rather than 3DGS;
(2)~it is the first to provide on-device measurement
comparisons across the four mobile / edge-GPU works
(\S\ref{sec:rendering}); and (3)~it is the first to combine
academic depth with the engineering roadmap of a single
mobile-SoC target (Snapdragon~8~Elite / Adreno~830 /
Vulkan~1.3). A second strand of related work is the
EG~2005 STAR ``Image-based Representations for Accelerated
Rendering of Complex Scenes'' by Jeschke et al., which
established the ``acquisition / representation / rendering''
three-tier structure that we adopt (with a four-tier extension
for the training-compression stage) in this survey.

\bigskip
\begin{itemize}
\item \textbf{2023-kerbl-3dgs}~\cite{kerbl20233dgs}: How to use
      anisotropic 3D Gaussians as a radiance-field representation,
      replacing NeRF's volumetric rendering with pure splatting.
\end{itemize}
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S A / B.
