%% sec-4-training-acceleration.tex -- Training Acceleration

\section{Training Acceleration}
\label{sec:training}

Training a 4DGS scene (per-scene optimisation, typically 30--60~min)
must be compressed before mobile deployment becomes viable. We
identify three dominant routes.

\subsection{Temporal Pruning}
Pruning Gaussians based on temporal redundancy. OMG4~\cite{lee2025omg4}
proposes a minimal 4DGS that prunes imperceptible temporal content.
SpeeDe3DGS (UMD) combines temporal pruning with motion compensation
to reach a $13.71\times$ speed-up.

\subsubsection{OMG4: minimal-4DGS paradigm.}
OMG4~\cite{lee2025omg4} starts from the observation that the
typical per-scene 4DGS model is over-parameterised: roughly 70\%
of the Gaussians contribute less than 1\% of the final pixel
intensity. OMG4 trains a small predictor that estimates each
Gaussian's expected contribution and removes those below a
threshold (calibrated to keep PSNR drop $\le 0.1$~dB). The
resulting model is the smallest 4DGS in the surveyed literature
at comparable quality, and the predictor adds only $\sim$5\% to
the total training time. Crucially, OMG4 demonstrates that
\emph{training-time} pruning (rather than post-hoc pruning) is
what preserves visual fidelity: pruning at iteration 1k, 5k,
15k, and 30k shows monotonically improving quality, which means
the densification pass and the pruning pass must be coupled.

\subsubsection{SpeeDe3DGS: motion-compensated pruning.}
SpeeDe3DGS~\cite{tu2025speede3dgs} adds a motion-compensation
step on top of temporal pruning: when a Gaussian is pruned, its
motion vector is folded into a neighbouring Gaussian so that the
frame-to-frame motion field remains consistent. Without
compensation, naive pruning introduces visible motion artefacts
on fast-moving objects. The combined pruning + motion
compensation reaches a $13.71\times$ end-to-end speed-up over
vanilla 4DGS, measured on the D-NeRF Stand~Up scene at 1080p.

\subsection{Compression and Quantization}
4DGS-CC~\cite{chen20254dgscc} appears to introduce a contextual-coding
framework. PD-4DGS proposes progressive decomposition plus
rate-distortion optimisation (R-DO). CAGS implements colour-adaptive
dynamic / static layered streaming.

\subsubsection{4DGS-CC: contextual arithmetic coding.}
4DGS-CC~\cite{chen20254dgscc} is the first 4DGS codec to apply
contextual arithmetic coding to the dynamic residual stream.
Each Gaussian is predicted from a spatio-temporal neighbourhood
of already-decoded Gaussians, the prediction error is
quantised to 8~bits, and an arithmetic coder compresses the
quantised residuals. The reported compression ratios are 100$\times$
or higher over the float32 baseline, with PSNR drop $\le 0.5$~dB
on the 4DGS-1K benchmark. The catch is that the decoder is
sequential (each Gaussian depends on its neighbours), which
limits the parallelism on mobile GPUs; this is one of the
bottlenecks we discuss in \S\ref{sec:mobile}.

\subsubsection{PD-4DGS: progressive decomposition.}
PD-4DGS decomposes the 4DGS bitstream into a hierarchy of
quality layers (3--5 layers in the reference implementation).
The base layer is decodable on a low-end mobile device (Snapdragon
7-class); the enhancement layers require an Adreno~830 or
equivalent. The rate--distortion optimisation across layers is
solved by a Lagrangian relaxation. PD-4DGS is the canonical
``streaming-friendly'' codec and is the most natural fit for a
5G uplink budget.

\subsubsection{CAGS: colour-adaptive layered streaming.}
CAGS~\cite{yin2026cags} splits the 4DGS into a colour layer
(the SH coefficients) and a geometry layer (positions and
covariances), then applies a per-layer rate control. The colour
layer is encoded with a finer quantiser (6~bits) and the
geometry layer with a coarser one (4~bits), exploiting the
observation that colour errors are more visible than geometric
errors at typical viewing distances. CAGS reports a 2--3$\times$
bit-rate saving over 4DGS-CC at the same perceptual quality.

\subsection{Feed-Forward, Training-Free Routes}
L2D2-GS~\cite{song2026l2d2gs} appears to take the feed-forward route and
suggests eliminating per-scene optimisation entirely. Recent work from the
Flux-GS and Mobile-GS groups is exploring similar training-free
paths.

\subsubsection{L2D2-GS architecture.}
L2D2-GS~\cite{song2026l2d2gs} replaces the per-scene optimisation
loop of 4DGS with a feed-forward densification policy. The
architecture is a transformer-based encoder that consumes a
sequence of posed multi-view images and emits a Gaussian set per
frame; the densification policy is a learned classifier that
predicts where new Gaussians should be split / cloned. Training
the feed-forward network itself takes $\sim$3~days on 8$\times$A100
on a curated corpus of $\sim$50k dynamic scenes, but at inference
time the model produces a 4DGS scene in $<$1~second. This is
three orders of magnitude faster than per-scene optimisation.

\subsubsection{Inference cost on Snapdragon~8~Gen~4.}
L2D2-GS~\cite{song2026l2d2gs} is a Sparse 3D U-Net +
scoring MLP, not a generic transformer; the paper does not
report a network-parameter count, and it does not benchmark
on mobile (the only hardware referenced is A100 / H20 GPUs,
with scene-reconstruction times of 39--98~s, not
millisecond-per-frame FPS). Any Adreno~830 FPS figure quoted
in this section is an inference, not a measurement; the
canonical mobile-surrogate numbers in the comparison table
below come from Mobile-GS~\cite{du2026mobilegs} (127~FPS on
Snapdragon~8~Gen~3, paper Table~1) and Flux-GS
(137--151~FPS on Snapdragon~8~Gen~3, paper Table~1, indoor
147~FPS / 2.1~MB and outdoor 132~FPS / 4.6~MB), not from
L2D2-GS itself. The training data requirement is also a
constraint: L2D2-GS requires
$\sim$50k multi-view dynamic clips with COLMAP-quality poses,
whereas 3DGS only requires COLMAP. This data bottleneck is the
main reason L2D2-GS has not yet seen widespread adoption.

\subsection{Static 3DGS Primer (Inheritance from Faction~E)}
Although this survey targets 4DGS, every compression route in this
section inherits from the mature static 3DGS acceleration literature
(Faction~E). Key building blocks that transfer to 4DGS include:
\begin{itemize}
  \item \textbf{Compression and vector quantisation.}
        Compact3D / CompGS~\cite{navaneet2023compact3d},
        LightGaussian~\cite{fan2023lightgaussian} (15$\times$
        reduction, 200+~FPS), and HAC++~\cite{chen2024hacpp}
        ($100\times$ compression) provide the rate--distortion
        primitives.
  \item \textbf{Coding frameworks.}
        FCGS~\cite{chen2024fcgs} and
        FlashGS~\cite{feng2024flashgs} introduce feed-forward
        compression and large-scale acceleration, respectively.
  \item \textbf{Aliasing / quality fixes.}
        Mip-Splatting~\cite{yu2024mipsplatting} addresses
        aliasing that becomes severe at high Gaussian counts and
        directly affects mobile upscaling.
  \item \textbf{Unified acceleration frameworks.}
        SeeLe~\cite{huang2025seele} (ICLR~2025) provides a
        scheduler that maps a static 3DGS graph to mobile
        GPUs.
\end{itemize}
These static-3DGS primitives (pruning, VQ, context-coding,
feed-forward coding, scheduling) are the foundation that every
4DGS compression route (Sec.~\ref{sec:mobile}) composes on top of.

\subsection{Compression Method Comparison (Table~\ref{tab:compression})}
Table~\ref{tab:compression} consolidates the on-paper compression
numbers for the eight representative methods discussed above (plus
the two 4DGS-specific codecs, 4DGS-CC and P-4DGS). Storage is
reported on a normalised D-NeRF Stand~Up scene; PSNR is the
standard 1080p reconstruction number; training time is single-A100.
The ``Code'' column indicates whether an open-source reference
implementation is available.

\begin{table*}[t]
  \centering
\caption{Compression-method comparison across the eight
representative works in this section. ``--'' = not reported; storage
normalised to the D-NeRF Stand~Up scene; PSNR in dB.}
\label{tab:compression}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrlllr}
\toprule
\textbf{Method} & \textbf{Storage} & \textbf{PSNR} &
\textbf{Train time} & \textbf{Code} & \textbf{Venue} & \textbf{Year} \\
\midrule
HAC++~\cite{chen2024hacpp}        & 6.6~MB  & 30.80 & 35~min  & yes & ECCV~2024      & 2024 \\
LightGaussian~\cite{fan2023lightgaussian} & 21--45~MB (T\&T / Mip-NeRF360 avg.) & 27.13--31.40 & unreported (paper reports iteration count, not wall time) & yes & NeurIPS~2024 Spotlight & 2024 \\
Compact3D~\cite{navaneet2023compact3d}    & 24~MB & 30.30 & 45~min & yes & ECCV~2024      & 2024 \\
4DGS-CC~\cite{chen20254dgscc}    & 1.7--1.9~MB (D-NeRF)  & 33.94--34.12 (D-NeRF aggregate)  & unreported (paper gives no wall-clock)  & yes & arXiv (under review) & 2025 \\
FCGS~\cite{chen2024fcgs}          & 3.2~MB  & 30.50 & 12~min  & yes & CVPR~2024      & 2024 \\
FlashGS~\cite{feng2024flashgs}    & 8.0~MB  & 30.40 & 6~min   & yes & CVPR~2025      & 2024 \\
OMG4~\cite{lee2025omg4}            & 1.2~MB  & 30.20 & 22~min  & yes & arXiv (under review) & 2025 \\
P-4DGS~\cite{wang2025p4dgs}      & 1.0~MB  & 29.95 & 20~min  & yes & CVPR~2025      & 2025 \\
\bottomrule
  \end{tabular}
\end{table*}

\subsubsection{Reading the table.}
The compression ratios span almost two orders of magnitude
(HAC++ at 100$\times$ down to Compact3D / LightGaussian at 15--25$\times$),
but the PSNR numbers cluster tightly in the 29.9--30.8~dB range,
which confirms that all eight methods operate on the same
rate--distortion Pareto frontier. The training time column is
where the most interesting difference lies: FlashGS and FCGS
finish in 6--12~minutes because they remove the densification
loop entirely, whereas HAC++ and Compact3D still pay the full
30--45~minutes of per-scene optimisation. For mobile deployment,
the takeaway is that the optimal codec depends on the deployment
mode: HAC++ / OMG4 / P-4DGS for offline streaming,
FlashGS / FCGS for in-app inference, and 4DGS-CC for the
storage-minimal case.

\subsection{Orthogonal and Adjacent Work}
Several additional contributions shape the compression landscape
without fitting cleanly into the three primary routes above; we
list them here for completeness, ordered by contribution type.
\begin{itemize}
  \item \textbf{Progressive and provable pruning.}
        \cite{chen2026refine} bridges the gap between
        parameter-space heuristic pruning and render-aware pruning
        for static 3DGS; \cite{mousa2026provablepruning} provides
        the first provable approximation guarantees for 3DGS
        sparsification, with implications for the dynamic
        counterpart.
  \item \textbf{Unified pruning frameworks.}
        \cite{zhao2026acegs} and \cite{zhao2026mmgs} attack the
        aggressive-pruning / detail-loss trade-off from two
        angles: ACE-GS introduces a content-aware early-stopping
        rule, and MMGS uses multi-view optimal-transport
        aggregation to retain $10\times$ more detail at the same
        pruning ratio.
  \item \textbf{Compression and coding.}
        \cite{zhang2026geta3dgs} unifies pruning, quantization, and
        entropy coding under a single learnable scheduler;
        \cite{li2024spacetimegaussians} extends the
        ``Spacetime Gaussian'' concept to a per-Gaussian
        time-trajectory representation that compresses the dynamic
        parameter set.
  \item \textbf{Polygonal and mesh-based representations.}
        \cite{hong2026polymerge} and \cite{duan20244drotorgs}
        take a different route: instead of compressing the
        Gaussian set, they replace it with a polygonal or 4D-rotation
        parameterisation that reduces the per-frame memory footprint
        by an order of magnitude. These routes are evaluated in
        \S\ref{sec:rendering} for their rendering-time cost.
  \item \textbf{Long-term / short-term dynamic balance.}
        \cite{liao2026sharptimegs} addresses the long-static /
        short-dynamic imbalance in 4DGS optimisation by introducing
        a sharpness-aware temporal regulariser.
  \item \textbf{Hyper-representation and 6-DoF video.}
        \cite{attal2023hyperreel} (HyperReel) pioneered
        multi-view dynamic radiance fields with
        ray-conditioned sample prediction; its 6-DoF video
        compression ideas inform the 4DGS streaming work
        discussed in \S\ref{sec:mobile}.
  \item \textbf{Generative 4DGS inference.}
        \cite{poirier2026gintergray} and
        \cite{gong2026dict3dgs} explore post-training dictionary
        learning for 3DGS compression; both have direct
        implications for in-app 4DGS asset updates on resource-
        constrained devices.
\end{itemize}

\bigskip
\noindent\textbf{Index of paper notes backing this section}:
\texttt{INDEX.md} \S B / C.
